### 2015-01-29 ggv.tex
init manuscript tex

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\usepackage{subcaption}
\usepackage{authblk}
\makeatletter
\renewcommand*{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or *\or**\or\dagger\or \ddagger\or
   \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger
   \or \ddagger\ddagger \else\@ctrerr\fi}}
\makeatother

\frenchspacing
\title{Visualizing the Geography of Genetic Variants}
\author[1]{Joseph Marcus$^{*}$}
\author[1]{John Novembre\thanks{Address correspondence to JM (\texttt{josephhmarcus@gmail.com}) or JN (\texttt{jnovembre@uchicago.edu}).}}
\affil[1]{\small{Department of Human Genetics, University of Chicago, Chicago, IL,USA}}

\date{}
\begin{document}
\maketitle

\begin{abstract}
One of the core features of any genetic variant, beyond its potential phenotypic effects or its frequency, is its geographic distribution. The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it survived and spread within, and in turn help us learn about historical patterns of migration and natural selection. Collectively the geographic distribution of genetic variants can help to explain how populations have been related through time (e.g. levels of gene flow and divergence). For variants with large effects, it can also help us understand the geographic distribution of spatially-varying phenotypes. For these reasons, visual inspection of geographic maps for genetic variants is common practice in genetic studies. Here we develop a series of reusable interactive visualizations for illuminating the geographic distribution of genetic variants. We specifically address several non-trivial challenges of this type of visualization; in particular, how to represent non-uniform levels of uncertainty in allele frequencies due to variable sample sizes; how to represent results from data with over 10,000 individuals in which allele frequencies can vary over 4 orders of magnitude; how to display data for regions of the globe with dense sampling of populations; and how to quickly access frequency data from large samples. To meet these challenges, we implement a flexible REST API for allowing for easy access to allele frequency and sample size data from large scale public genomic datasets. Built upon this API we develop a web-based browser, entitled the Geography of Genetic Variants (GGV) browser for visualizing the geographic distribution of genetic variants. The GGV browser rapidly provides maps of derived allele frequencies in populations distributed across the globe. The GGV browser builds upon past tools such as the HGDP Selection browser by allowing for more interactive features, new representations of rare variation, as well as incorporating uncertainty in allele frequency estimation. As ancillaries, we also develop a research visualization toolkit that includes a method for displaying high Fst outlier SNPs from the joint site frequency spectrum and an interactive version of commonly used PCA figures. We hope the GGV browser will be a valuable research and education tool for exploring population genetics data.

\end{abstract}
\newpage

%%%%%%%% %%
\section*{Introduction}

\end{document}

### 2016-06-30 paper.Rmd
reformatting / structuring adding figures

---
title:
  # Title must be 150 characters or less
  "Visualizing the geography of genetic variants"
author:
  - name: Joseph H. Marcus
    affiliation: '1'
    # If corresponding author (only one corresponding author per manuscript)
    email: jhmarcus@uchicago.edu
  - name: John Novembre
    email: jnovembre@uchicago.edu
    affiliation: '1'
affiliation:
  - id: '1'
    department: Department of Human Genetics
    institution: University of Chicago
    city: Chicago
    state: IL
    country: USA
output: RmdTemplates::plos_article
csl: plos.csl
bibliography: library.bib
---

```{r imports, echo=FALSE, message=FALSE, warning=FALSE}
library(png)
library(grid)
```

# Abstract {-}
One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its frequency, is its geographic distribution. The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations spread within, and in turn can help one learn about historical patterns of migration, genetic drift, and natural selection. Collectively the geographic distribution of genetic variants can help to explain how populations have been related through time (e.g. levels of gene flow and divergence). For these reasons, visual inspection of geographic maps for genetic variants is common practice in genetic studies. Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants. Through an efficient RESTful API and dynamic front-end the Geography of Genetic Variants (GGV) browser rapidly provides maps of allele frequencies in populations distributed across the globe.

# Introduction {-}
Genetics researchers often face the following problem: the researcher has identified one or more genetic variants of interest (e.g. from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant. For example, the researcher may hope to address: 1) implications for genomic medicine (e.g. is a risk allele geographically localized to a certain patient population?) @XXX; 2) design of follow-up studies (e.g. what population should be studied to observe variant carriers?) @XXX; or 3) the biology of the variant in question (e.g. does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?) @novembre2009spatial.  A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.

Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets. For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide vairants (SNVs). One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast.   While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail. A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.

Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods @furchedata and geographic plotting techniques that are unfamiliar to most. Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.

Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.  One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a sbustantial learning curve.  The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @XXX, books e.g. @dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the XX track of the browser).

Reference datasets for population genetic variation has greatly expanded since the release of the HGDP Illumina 650Y dataset @li2008worldwide that formed the basis of the HGDP Selection Browser maps.  The most notable advance is the publication of the 1000 Genomes phase 3 data @10002015global though additional datasets are coming online @meyer2012high, @lazaridis2014ancient.

In addition, novel approaches for data visualization have become more widely available.  In particular, web-based visualization tools, such as Data Driven Documents (d3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.

Taking advantage of these recent advances, we aim to address significant visualization challenges that are inherit in the production of geographic allele frequency maps, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.

# Fundamental Approach {-}
The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of d3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.

In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project, Human Genome Diversity project and POPRES project, we have developed an SQL database and RESTful API for querying allele frequencies by chromosome and position or reference SNP identifier [Figure 1].

```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
img <- readPNG("figures/figure1.png")
grid.raster(img)
```

Users can query single variants by chromosome and position identifiers, by rsids @sherry2001dbsnp, or users can simply choose a random variant from within a dataset.  While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset. For instance, in data with deep population subdivision, it is obvious in the consistent patterns of differentiation observed across markers. We also expect this will be useful in teaching contexts – as it provides a highly visual way for learners to understand human genetic variation.

After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies. Pie charts are displayed as points positioned at the population's region of origin where the projection/map-view is chosen based off of the geographic proximity of populations present in a given dataset [Figure 2].

# Representing variable certainty in frequency data {-}
One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty. For some locations, sample sizes are small, and the reported allele frequency may be quite far from the true population frequency due to sampling error. To address this issue, we use varying transparency in a population's pie chart: sample allele frequencies with higher levels of sampling error will be made more transparent, and hence less visible on the map [Figure 3].

# Representing rare variants in frequency data {-}
An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset. This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays biased towards variants that are common in human populations (5-50% in minor allele frequency). With the advent of next generation sequencing technologies and large samples of thousands of individuals, it is common for datasets to contain rare variants @10002015global, @nelson2012abundance, @tennessen2012evolution.

In current visualization schemes, such as the HGDP Selection Browser, rare variants would be represented as narrow slivers in a pie chart, nearly invisible to the naked eye. To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible. Specifically, we will use a “frequency scale” that is indicated below the map, and redundantly using color, that will indicate a constant scale for the frequencies displayed [Figure 4]. Much like scientific notation, this allows a wide range of frequencies to be displayed].

# Interface features {-}
In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information. To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins. Lines anchoring the points visibly to their original sampling locations will be used to make sure true sampling locations are indicated [Figure 5].

# Underlying Frequency API {-}
[Joe - worth saying more about the API structure and giving example calls and structure of JSON?]

# Caveats {-}
A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.  While doing so is generally immensely helpful, it has inherent complexity.  For example,  pracitioners must make choices regarding representing where an individual was sampled for the study (e.g. the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g. as determined in practice by the birthplaces of recent ancestors, such as grandparents).  We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that aligns with those used by the initial analysts of the data.  A future feature will be to allow alternative location schema to be used for the populations in a dataset.

# Conclusion {-}
By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.  We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.

# Acknowledgements {-}
Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank XX for supportive conversations.

# Acknowledgements {-}

# References {-}

### 2016-06-30 paper.tex
reformatting / structuring adding figures

% Template for PLoS

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Visualizing the geography of genetic variants}
}
% Insert Author names, affiliations and corresponding author email.
\\
  Joseph H. Marcus\textsuperscript{1*}, 
  John Novembre\textsuperscript{1*}\\
\bf{1} Department of Human Genetics, University of Chicago,  Chicago,  IL,  USA
\\

\textasteriskcentered{} E-mail:   \href{mailto:jhmarcus@uchicago.edu}{\nolinkurl{jhmarcus@uchicago.edu}}
  \href{mailto:jnovembre@uchicago.edu}{\nolinkurl{jnovembre@uchicago.edu}}

\end{flushleft}

\section*{Abstract}\label{abstract}
\addcontentsline{toc}{section}{Abstract}

One of the key characteristics of any genetic variant, beyond its
potential phenotypic effects or its frequency, is its geographic
distribution. The geographic distribution of a genetic variant can shed
light on where the variant first arose, in what populations spread
within, and in turn can help one learn about historical patterns of
migration, genetic drift, and natural selection. Collectively the
geographic distribution of genetic variants can help to explain how
populations have been related through time (e.g.~levels of gene flow and
divergence). For these reasons, visual inspection of geographic maps for
genetic variants is common practice in genetic studies. Here we develop
an interactive visualization tool for illuminating the geographic
distribution of genetic variants. Through an efficient RESTful API and
dynamic front-end the Geography of Genetic Variants (GGV) browser
rapidly provides maps of allele frequencies in populations distributed
across the globe.

\section*{Introduction}\label{introduction}
\addcontentsline{toc}{section}{Introduction}

Genetics researchers often face the following problem: the researcher
has identified one or more genetic variants of interest (e.g.~from a
genome-wide association, eQTL, or pharmacogenomic study) and would like
to know the geographic distribution of the variant. For example, the
researcher may hope to address: 1) implications for genomic medicine
(e.g.~is a risk allele geographically localized to a certain patient
population?) \textbf{???}; 2) design of follow-up studies (e.g.~what
population should be studied to observe variant carriers?) \textbf{???};
or 3) the biology of the variant in question (e.g.~does the variant
correlate with a known environmental factor in a manner suggestive of
some geographically localized selection pressure?) {[}1{]}. A simple
geographic map of the distribution of a genetic variant can be
incredibly insightful for these questions.

Contemporary population genetics researchers are also faced with the
challenge of large, high dimensional datasets. For example, is not
uncommon for a researcher in human genetics to have a dataset comprised
of thousands of individuals measured at hundreds of thousands or even
millions of single nucleotide vairants (SNVs). One common approach to
visualizing such high-dimensional data is to compress the SNV dimensions
down to a small number of latent factors, using a method such as
principal components analysis {[}2{]}, {[}3{]}, a model-based clustering
method such as STRUCTURE {[}4{]} or ADMIXTURE {[}5{]}. While these
approaches are extremely valuable, researchers can use them to often
without inspecting the underlying variant data in more detail. A natural
approach to gaining more insight to the overall structure of a
population genetic dataset is to visually inspect what geographic
patterns arise in allele frequency maps.

Unfortunately, generating geographic allele frequency maps is
time-consuming for the average researcher as it requires a combination
of data wrangling methods {[}6{]} and geographic plotting techniques
that are unfamiliar to most. Our aim here is to produce a tailored
system for rapidly constructing informative geographic maps of allele
frequency variation.

Our work is inspired by past tools such as the ALFRED database {[}7{]}
and the maps available on the HGDP Selection browser {[}8{]}. One of us
(J.N.) developed the scripts for the HGDP Selection Browser maps using
The Generic Mapping Tools (GMT) {[}9{]}, a powerful system of geographic
plotting scripts with a sbustantial learning curve. The plots from the
HGDP Selection Browser have proved useful, and have appeared in research
articles e.g \textbf{???}, books e.g. {[}10{]}, and have been ported and
made avialable on the UCSC Genome Browser (available under the XX track
of the browser).

Reference datasets for population genetic variation has greatly expanded
since the release of the HGDP Illumina 650Y dataset {[}11{]} that formed
the basis of the HGDP Selection Browser maps. The most notable advance
is the publication of the 1000 Genomes phase 3 data {[}12{]} though
additional datasets are coming online {[}13{]}, {[}14{]}.

In addition, novel approaches for data visualization have become more
widely available. In particular, web-based visualization tools, such as
Data Driven Documents (d3.js), offer powerful approaches to realize
these aims through useful methods for interactivity, advantages of
software development in modern web-browsers, a large open-source
development community, and ease of share-ability {[}15{]}.

Taking advantage of these recent advances, we aim to address significant
visualization challenges that are inherit in the production of
geographic allele frequency maps, including dynamic interaction, display
of rare genetic variation, and representation of uncertainty in
estimated allele frequencies due to variable sample sizes.

\section*{Fundamental Approach}\label{fundamental-approach}
\addcontentsline{toc}{section}{Fundamental Approach}

The Geography of Genetic Variants browser (GGV) uses the SVG and mapping
utilities of d3.js to generate interactive frequency maps, allowing for
quick and dynamic displays of the geographic distribution of a genetic
variant.

In order to allow for easy access to large commonly used public genomic
datasets, such as the 1000 Genomes project, Human Genome Diversity
project and POPRES project, we have developed an SQL database and
RESTful API for querying allele frequencies by chromosome and position
or reference SNP identifier {[}Figure 1{]}.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/figure1-1}

}

\caption[Figure caption]{Figure caption}\label{fig:figure1}
\end{figure}
\end{CodeChunk}

Users can query single variants by chromosome and position identifiers,
by rsids {[}16{]}, or users can simply choose a random variant from
within a dataset. While many applications require inspection of the
distribution of a specific variant, from our experience, it can be very
helpful to quickly view the geographic distribution of several randomly
chosen variants to quickly gain a sense of structure in a dataset. For
instance, in data with deep population subdivision, it is obvious in the
consistent patterns of differentiation observed across markers. We also
expect this will be useful in teaching contexts -- as it provides a
highly visual way for learners to understand human genetic variation.

After a query, the GGV displays the frequency of a variant in a given
population as a pie chart where each slice represents minor and major
allele frequencies. Pie charts are displayed as points positioned at the
population's region of origin where the projection/map-view is chosen
based off of the geographic proximity of populations present in a given
dataset {[}Figure 2{]}.

\section*{Representing variable certainty in frequency
data}\label{representing-variable-certainty-in-frequency-data}
\addcontentsline{toc}{section}{Representing variable certainty in
frequency data}

One under-appreciated problem with allele frequency maps is that not all
data points have equal levels of certainty. For some locations, sample
sizes are small, and the reported allele frequency may be quite far from
the true population frequency due to sampling error. To address this
issue, we use varying transparency in a population's pie chart: sample
allele frequencies with higher levels of sampling error will be made
more transparent, and hence less visible on the map {[}Figure 3{]}.

\section*{Representing rare variants in frequency
data}\label{representing-rare-variants-in-frequency-data}
\addcontentsline{toc}{section}{Representing rare variants in frequency
data}

An additional challenge is that allele frequencies between variants
often differ greatly, sometimes by orders of magnitude in a single
dataset. This has not been an pervasive problem until recently, as most
population genetic samples were based on genotype arrays biased towards
variants that are common in human populations (5-50\% in minor allele
frequency). With the advent of next generation sequencing technologies
and large samples of thousands of individuals, it is common for datasets
to contain rare variants {[}12{]}, {[}17{]}, {[}18{]}.

In current visualization schemes, such as the HGDP Selection Browser,
rare variants would be represented as narrow slivers in a pie chart,
nearly invisible to the naked eye. To address this challenge we re-scale
frequencies for rare variants, so that small frequencies become visible.
Specifically, we will use a ``frequency scale'' that is indicated below
the map, and redundantly using color, that will indicate a constant
scale for the frequencies displayed {[}Figure 4{]}. Much like scientific
notation, this allows a wide range of frequencies to be displayed{]}.

\section*{Interface features}\label{interface-features}
\addcontentsline{toc}{section}{Interface features}

In many datasets where populations are sampled densely in geographic
space, one problem is that allele frequency plots begin to overlap each
other and obscure information. To address this issue, we use
force-directed layouts of the populations such that no two points are
overlapping each other, and yet the points will be pulled towards their
true origins. Lines anchoring the points visibly to their original
sampling locations will be used to make sure true sampling locations are
indicated {[}Figure 5{]}.

\section*{Underlying Frequency API}\label{underlying-frequency-api}
\addcontentsline{toc}{section}{Underlying Frequency API}

{[}Joe - worth saying more about the API structure and giving example
calls and structure of JSON?{]}

\section*{Caveats}\label{caveats}
\addcontentsline{toc}{section}{Caveats}

A major challenge of using a geographic representation of genetic
variation in humans is that the samples must be associated with a
geographic location. While doing so is generally immensely helpful, it
has inherent complexity. For example, pracitioners must make choices
regarding representing where an individual was sampled for the study
(e.g.~the city of a major research center) or choosing a location that
is more representative of an individual's ancestral origins (e.g.~as
determined in practice by the birthplaces of recent ancestors, such as
grandparents). We do not proscribe a general solution to this problem
and instead use a set of locations for each dataset that aligns with
those used by the initial analysts of the data. A future feature will be
to allow alternative location schema to be used for the populations in a
dataset.

\section*{Conclusion}\label{conclusion}
\addcontentsline{toc}{section}{Conclusion}

By allowing rapid generation of allele frequency maps, we hope to
facilitate the interpretation of variant function and history by
practicing geneticists. We also hope the ability to query random
variants from major human population genetic samples will allow students
to appreciate the structure of human genetic diversity in a more
approachable and intuitive form than alternative visualizations.

\section*{Acknowledgements}\label{acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

Support for this work was provided by the NIH-BD2K initiative (1U01
CA198933-0). The authors would also like to thank XX for supportive
conversations.

\section*{Acknowledgements}\label{acknowledgements-1}
\addcontentsline{toc}{section}{Acknowledgements}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-novembre2009spatial}{}
1. Novembre J, Di Rienzo A. Spatial patterns of variation due to natural
selection in humans. Nature Reviews Genetics. Nature Publishing Group;
2009;10: 745--755.

\hypertarget{ref-price2006principal}{}
2. Price AL, Patterson NJ, Plenge RM, Weinblatt ME, Shadick NA, Reich D.
Principal components analysis corrects for stratification in genome-wide
association studies. Nature genetics. Nature Publishing Group; 2006;38:
904--909.

\hypertarget{ref-patterson2006population}{}
3. Patterson N, Price AL, Reich D. Population structure and
eigenanalysis. PLoS genet. Public Library of Science; 2006;2: e190.

\hypertarget{ref-pritchard2000inference}{}
4. Pritchard JK, Stephens M, Donnelly P. Inference of population
structure using multilocus genotype data. Genetics. Genetics Soc
America; 2000;155: 945--959.

\hypertarget{ref-alexander2009fast}{}
5. Alexander DH, Novembre J, Lange K. Fast model-based estimation of
ancestry in unrelated individuals. Genome research. Cold Spring Harbor
Lab; 2009;19: 1655--1664.

\hypertarget{ref-furchedata}{}
6. Furche T, Gottlob G, Libkin L, Orsi G, Paton NW. Data wrangling for
big data: Challenges and opportunities.

\hypertarget{ref-rajeevan2011alfred}{}
7. Rajeevan H, Soundararajan U, Kidd JR, Pakstis AJ, Kidd KK. ALFRED: An
allele frequency resource for research and teaching. Nucleic acids
research. Oxford Univ Press; 2011; gkr924.

\hypertarget{ref-pickrell2009signals}{}
8. Pickrell JK, Coop G, Novembre J, Kudaravalli S, Li JZ, Absher D, et
al. Signals of recent positive selection in a worldwide sample of human
populations. Genome research. Cold Spring Harbor Lab; 2009;19: 826--837.

\hypertarget{ref-wessel2013generic}{}
9. Wessel P, Smith WH, Scharroo R, Luis J, Wobbe F. Generic mapping
tools: Improved version released. Eos, Transactions American Geophysical
Union. Wiley Online Library; 2013;94: 409--410.

\hypertarget{ref-dudley2013exploring}{}
10. Dudley JT, Karczewski KJ. Exploring personal genomics. Oxford
University Press; 2013.

\hypertarget{ref-li2008worldwide}{}
11. Li JZ, Absher DM, Tang H, Southwick AM, Casto AM, Ramachandran S, et
al. Worldwide human relationships inferred from genome-wide patterns of
variation. science. American Association for the Advancement of Science;
2008;319: 1100--1104.

\hypertarget{ref-10002015global}{}
12. Consortium 1GP, others. A global reference for human genetic
variation. Nature. Nature Publishing Group; 2015;526: 68--74.

\hypertarget{ref-meyer2012high}{}
13. Meyer M, Kircher M, Gansauge M-T, Li H, Racimo F, Mallick S, et al.
A high-coverage genome sequence from an archaic denisovan individual.
Science. American Association for the Advancement of Science; 2012;338:
222--226.

\hypertarget{ref-lazaridis2014ancient}{}
14. Lazaridis I, Patterson N, Mittnik A, Renaud G, Mallick S, Kirsanow
K, et al. Ancient human genomes suggest three ancestral populations for
present-day europeans. Nature. Nature Publishing Group; 2014;513:
409--413.

\hypertarget{ref-bostock2011d3}{}
15. Bostock M, Ogievetsky V, Heer J. D\(^3\) data-driven documents.
Visualization and Computer Graphics, IEEE Transactions on. IEEE;
2011;17: 2301--2309.

\hypertarget{ref-sherry2001dbsnp}{}
16. Sherry ST, Ward M-H, Kholodov M, Baker J, Phan L, Smigielski EM, et
al. DbSNP: The nCBI database of genetic variation. Nucleic acids
research. Oxford Univ Press; 2001;29: 308--311.

\hypertarget{ref-nelson2012abundance}{}
17. Nelson MR, Wegmann D, Ehm MG, Kessner D, Jean PS, Verzilli C, et al.
An abundance of rare functional variants in 202 drug target genes
sequenced in 14,002 people. Science. American Association for the
Advancement of Science; 2012;337: 100--104.

\hypertarget{ref-tennessen2012evolution}{}
18. Tennessen JA, Bigham AW, O'Connor TD, Fu W, Kenny EE, Gravel S, et
al. Evolution and functional impact of rare coding variation from deep
sequencing of human exomes. Science. American Association for the
Advancement of Science; 2012;337: 64--69.

\end{document}

### 2016-06-30 paper.Rmd
adding figures

@@
 ```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
-img <- readPNG("figures/figure1.png")
+img <- readPNG("figures/png/figure1.png")
 grid.raster(img)
@@
 
+```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
+img <- readPNG("figures/png/figure2.png")
+grid.raster(img)
+```
+
 # Representing variable certainty in frequency data {-}
@@
 To address this issue, we use varying transparency in a population's pie chart: sample allele frequencies with higher levels of sampling error will be made more transparent, and hence less visible on the map [Figure 3].
+
+```{r figure3, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Figure caption'}
+img <- readPNG("figures/png/figure3.png")
+grid.raster(img)
+```
 
@@
 
+```{r figure4, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Figure caption'}
+img <- readPNG("figures/png/figure4.png")
+grid.raster(img)
+```
+
 # Interface features {-}
@@
 Lines anchoring the points visibly to their original sampling locations will be used to make sure true sampling locations are indicated [Figure 5].
+
+```{r figure5, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Figure caption'}
+img <- readPNG("figures/png/figure5.png")
+grid.raster(img)
+```
 

### 2016-06-30 paper.Rmd
added more text, nearly draft finished figures and started fig captions

@@
 One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its frequency, is its geographic distribution.
-The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations spread within, and in turn can help one learn about historical patterns of migration, genetic drift, and natural selection.
+The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it spread within, and in turn can help one learn about historical patterns of migration, genetic drift, and natural selection.
 Collectively the geographic distribution of genetic variants can help to explain how populations have been related through time (e.g.
@@
 Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants.
-Through an efficient RESTful API and dynamic front-end the Geography of Genetic Variants (GGV) browser rapidly provides maps of allele frequencies in populations distributed across the globe.
+Through an efficient REST API and dynamic front-end the Geography of Genetic Variants (GGV) browser rapidly provides maps of allele frequencies in populations distributed across the globe.
 
 # Introduction {-}
-Genetics researchers often face the following problem: the researcher has identified one or more genetic variants of interest (e.g.
+Genetics researchers often face the following problem: the researcher has identified one or many genetic variants of interest (e.g.
 from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant.
 For example, the researcher may hope to address: 1) implications for genomic medicine (e.g.
-is a risk allele geographically localized to a certain patient population?) @XXX; 2) design of follow-up studies (e.g.
-what population should be studied to observe variant carriers?) @XXX; or 3) the biology of the variant in question (e.g.
-does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?) @novembre2009spatial.
+is a risk allele geographically localized to a certain patient population?); 2) design of follow-up studies (e.g.
+what population should be studied to observe variant carriers?); or 3) the biology of the variant in question (e.g.
+does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?)  @rosenberg2010genome, @novembre2009spatial, @coop2010using.
 A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
+
+The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry.
+Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration @nelson2012abundance, @mathieson2014demography.
+Thus by surveying the collection of rare variants that an individual carries provides information about the population origins of that individual's recent ancestors.
+One of us (J.N.) recently took advantage of the geographic clustering of an individual's rare variants in helping to reaffirm the historically highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.
 
@@
 One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a sbustantial learning curve.
-The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @XXX, books e.g.
-@dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the XX track of the browser).
+The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @pickrell2009signals, @coop2009role, books e.g.
+@dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
 
@@
 In addition, novel approaches for data visualization have become more widely available.
-In particular, web-based visualization tools, such as Data Driven Documents (d3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.
+In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.
 
@@
 # Fundamental Approach {-}
-The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of d3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
 
-In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project, Human Genome Diversity project and POPRES project, we have developed an SQL database and RESTful API for querying allele frequencies by chromosome and position or reference SNP identifier [Figure 1].
+The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of D3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
 
-```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
+```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Schematic diagram of the Geography of Genetic Variants (GGV) web application.
+We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genom Diversity Project (HGDP) and Population Reference Sample (POPRES).
+These allele freqeuncies are stored in a SQL database along side with geographic coordinates associated with each population.
+The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.'}
 img <- readPNG("figures/png/figure1.png")
@@
 
-Users can query single variants by chromosome and position identifiers, by rsids @sherry2001dbsnp, or users can simply choose a random variant from within a dataset.
+In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project, Human Genome Diversity project and POPRES project, we have developed an SQL database and REST API for querying allele frequencies by chromosome and position or reference SNP identifier [Figure 1].
+
+Users can query single variants by chromosome and position identifiers, by rsid @sherry2001dbsnp, or users can simply choose a random variant from within a dataset.
 While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
@@
 
-After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies.
-Pie charts are displayed as points positioned at the population's region of origin where the projection/map-view is chosen based off of the geographic proximity of populations present in a given dataset [Figure 2].
-
-```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
+```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application.
+Each pie chart represents a population with the blue slice of the pie displaying the freqeuncy of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.'}
 img <- readPNG("figures/png/figure2.png")
@@
 ```
+
+After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies.
+Pie charts are displayed as points positioned at the population's region of origin where the projection / map-view is chosen based off of the geographic proximity of populations present in a given dataset [Figure 2].
 
@@
 An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset.
-This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays biased towards variants that are common in human populations (5-50% in minor allele frequency).
+This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays, biased towards variants that are common in human populations (5-50% in minor allele frequency).
 With the advent of next generation sequencing technologies and large samples of thousands of individuals, it is common for datasets to contain rare variants @10002015global, @nelson2012abundance, @tennessen2012evolution.
@@
 # Underlying Frequency API {-}
-[Joe - worth saying more about the API structure and giving example calls and structure of JSON?]
+To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API impelemented in the python library Flask-RESTful @citeXXX.
+GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g.
+latitude, longitude, population label, sample size, and frequency scale.
+Genetic variants can be queried by chromosome position, rsid, or randomly.
+Example HTTP requests and json response can be seen below:
+
+## Query API by rsid {-}
+```
+http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&rsID=rs1834640
+[
+  {
+    "alleles": ["A", "G"], 
+    "pos": ["-15.310139", "13.443182"], 
+    "pop": "GWD", 
+    "nobs": "226", 
+    "xobs": "17", 
+    "freqscale": 1, 
+    "freq": [0.0752212389381, 0.9247787610619], 
+    "chrom_pos": "15:48392165", 
+    "rawfreq": 0.0752212389381
+  },
+  ...
+]
+```
+
+## Query API by chromosome position {-}
+
+```
+http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&chr=14&pos=37690093
+[
+  {
+    "alleles": ["G", "A"], 
+    "pos": ["-15.310139", "13.443182"], 
+    "pop": "GWD", 
+    "nobs": "226", 
+    "xobs": "0", 
+    "freqscale": 0.01, 
+    "freq": [0.0, 1.0], 
+    "chrom_pos": "14:37690093", 
+    "rawfreq": 0.0
+  }
+  ...
+]
+```
+
+## Randomly query API {-}
+
+```
+http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&random_snp=True
+[
+  {
+    "alleles": ["T", "C"], 
+    "pos": ["-15.310139", "13.443182"], 
+    "pop": "GWD", 
+    "nobs": "226", 
+    "xobs": "0", 
+    "freqscale": 0.01, 
+    "freq": [0.0, 1.0], 
+    "chrom_pos": "5:42452893", 
+    "rawfreq": 0.0
+  }, 
+  ...
+]
+```
 
@@
 # Acknowledgements {-}
-Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank XX for supportive conversations.
-
-# Acknowledgements {-}
+Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank members of the Novembre Lab for supportive conversations.
 

### 2016-06-30 paper.tex
adding figures
added more text, nearly draft finished figures and started fig captions

@@
 distribution. The geographic distribution of a genetic variant can shed
-light on where the variant first arose, in what populations spread
+light on where the variant first arose, in what populations it spread
 within, and in turn can help one learn about historical patterns of
@@
 an interactive visualization tool for illuminating the geographic
-distribution of genetic variants. Through an efficient RESTful API and
+distribution of genetic variants. Through an efficient REST API and
 dynamic front-end the Geography of Genetic Variants (GGV) browser
@@
 Genetics researchers often face the following problem: the researcher
-has identified one or more genetic variants of interest (e.g.~from a
+has identified one or many genetic variants of interest (e.g.~from a
 genome-wide association, eQTL, or pharmacogenomic study) and would like
@@
 (e.g.~is a risk allele geographically localized to a certain patient
-population?) \textbf{???}; 2) design of follow-up studies (e.g.~what
-population should be studied to observe variant carriers?) \textbf{???};
-or 3) the biology of the variant in question (e.g.~does the variant
-correlate with a known environmental factor in a manner suggestive of
-some geographically localized selection pressure?) {[}1{]}. A simple
+population?); 2) design of follow-up studies (e.g.~what population
+should be studied to observe variant carriers?); or 3) the biology of
+the variant in question (e.g.~does the variant correlate with a known
+environmental factor in a manner suggestive of some geographically
+localized selection pressure?) {[}1{]}, {[}2{]}, {[}3{]}. A simple
 geographic map of the distribution of a genetic variant can be
 incredibly insightful for these questions.
+
+The geographic distribution of rare genetic variants (minor allele
+frequency \(\leq .05\)) can be particularly informative in determining
+an individual's ancestry. Rare variants tend to be localized in small
+geographic regions because they have arisen recently in human history
+and have had little time to spread across populations through migration
+{[}4{]}, {[}5{]}. Thus by surveying the collection of rare variants that
+an individual carries provides information about the population origins
+of that individual's recent ancestors. One of us (J.N.) recently took
+advantage of the geographic clustering of an individual's rare variants
+in helping to reaffirm the historically highly contested Native American
+ancestry of the Kennewick Man, an ancient individual from Washington
+State \textbf{???}.
 
@@
 down to a small number of latent factors, using a method such as
-principal components analysis {[}2{]}, {[}3{]}, a model-based clustering
-method such as STRUCTURE {[}4{]} or ADMIXTURE {[}5{]}. While these
+principal components analysis {[}6{]}, {[}7{]}, a model-based clustering
+method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While these
 approaches are extremely valuable, researchers can use them to often
@@
 time-consuming for the average researcher as it requires a combination
-of data wrangling methods {[}6{]} and geographic plotting techniques
+of data wrangling methods {[}10{]} and geographic plotting techniques
 that are unfamiliar to most. Our aim here is to produce a tailored
@@
 
-Our work is inspired by past tools such as the ALFRED database {[}7{]}
-and the maps available on the HGDP Selection browser {[}8{]}. One of us
+Our work is inspired by past tools such as the ALFRED database {[}11{]}
+and the maps available on the HGDP Selection browser {[}12{]}. One of us
 (J.N.) developed the scripts for the HGDP Selection Browser maps using
-The Generic Mapping Tools (GMT) {[}9{]}, a powerful system of geographic
-plotting scripts with a sbustantial learning curve. The plots from the
-HGDP Selection Browser have proved useful, and have appeared in research
-articles e.g \textbf{???}, books e.g. {[}10{]}, and have been ported and
-made avialable on the UCSC Genome Browser (available under the XX track
-of the browser).
+The Generic Mapping Tools (GMT) {[}13{]}, a powerful system of
+geographic plotting scripts with a sbustantial learning curve. The plots
+from the HGDP Selection Browser have proved useful, and have appeared in
+research articles e.g {[}12{]}, {[}14{]}, books e.g. {[}15{]}, and have
+been ported and made avialable on the UCSC Genome Browser (available
+under the HGDP Allele Freq track of the browser).
 
 Reference datasets for population genetic variation has greatly expanded
-since the release of the HGDP Illumina 650Y dataset {[}11{]} that formed
+since the release of the HGDP Illumina 650Y dataset {[}16{]} that formed
 the basis of the HGDP Selection Browser maps. The most notable advance
-is the publication of the 1000 Genomes phase 3 data {[}12{]} though
-additional datasets are coming online {[}13{]}, {[}14{]}.
+is the publication of the 1000 Genomes phase 3 data {[}17{]} though
+additional datasets are coming online {[}18{]}, {[}19{]}.
 
@@
 widely available. In particular, web-based visualization tools, such as
-Data Driven Documents (d3.js), offer powerful approaches to realize
+Data Driven Documents (D3.js), offer powerful approaches to realize
 these aims through useful methods for interactivity, advantages of
 software development in modern web-browsers, a large open-source
-development community, and ease of share-ability {[}15{]}.
+development community, and ease of share-ability {[}20{]}.
 
@@
 The Geography of Genetic Variants browser (GGV) uses the SVG and mapping
-utilities of d3.js to generate interactive frequency maps, allowing for
+utilities of D3.js to generate interactive frequency maps, allowing for
 quick and dynamic displays of the geographic distribution of a genetic
 variant.
-
-In order to allow for easy access to large commonly used public genomic
-datasets, such as the 1000 Genomes project, Human Genome Diversity
-project and POPRES project, we have developed an SQL database and
-RESTful API for querying allele frequencies by chromosome and position
-or reference SNP identifier {[}Figure 1{]}.
 
@@
 
-\caption[Figure caption]{Figure caption}\label{fig:figure1}
+\caption[Schematic diagram of the Geography of Genetic Variants (GGV) web application]{Schematic diagram of the Geography of Genetic Variants (GGV) web application.
+We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genom Diversity Project (HGDP) and Population Reference Sample (POPRES).
+These allele freqeuncies are stored in a SQL database along side with geographic coordinates associated with each population.
+The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
 \end{figure}
@@
 
+In order to allow for easy access to large commonly used public genomic
+datasets, such as the 1000 Genomes project, Human Genome Diversity
+project and POPRES project, we have developed an SQL database and REST
+API for querying allele frequencies by chromosome and position or
+reference SNP identifier {[}Figure 1{]}.
+
 Users can query single variants by chromosome and position identifiers,
-by rsids {[}16{]}, or users can simply choose a random variant from
+by rsid {[}21{]}, or users can simply choose a random variant from
 within a dataset. While many applications require inspection of the
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/figure2-1}
+
+}
+
+\caption[Example map from the Geography of Genetic Variants web application]{Example map from the Geography of Genetic Variants web application.
+Each pie chart represents a population with the blue slice of the pie displaying the freqeuncy of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:figure2}
+\end{figure}
+\end{CodeChunk}
+
 After a query, the GGV displays the frequency of a variant in a given
@@
 allele frequencies. Pie charts are displayed as points positioned at the
-population's region of origin where the projection/map-view is chosen
+population's region of origin where the projection / map-view is chosen
 based off of the geographic proximity of populations present in a given
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/figure3-1}
+
+}
+
+\caption[Figure caption]{Figure caption}\label{fig:figure3}
+\end{figure}
+\end{CodeChunk}
+
 \section*{Representing rare variants in frequency
@@
 dataset. This has not been an pervasive problem until recently, as most
-population genetic samples were based on genotype arrays biased towards
+population genetic samples were based on genotype arrays, biased towards
 variants that are common in human populations (5-50\% in minor allele
@@
 and large samples of thousands of individuals, it is common for datasets
-to contain rare variants {[}12{]}, {[}17{]}, {[}18{]}.
+to contain rare variants {[}17{]}, {[}4{]}, {[}22{]}.
 
@@
 notation, this allows a wide range of frequencies to be displayed{]}.
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/figure4-1}
+
+}
+
+\caption[Figure caption]{Figure caption}\label{fig:figure4}
+\end{figure}
+\end{CodeChunk}
 
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/figure5-1}
+
+}
+
+\caption[Figure caption]{Figure caption}\label{fig:figure5}
+\end{figure}
+\end{CodeChunk}
+
 \section*{Underlying Frequency API}\label{underlying-frequency-api}
@@
 
-{[}Joe - worth saying more about the API structure and giving example
-calls and structure of JSON?{]}
+To interface population minor allele frequency data stored in the SQL
+database with the front-end D3.js visualization we use a REST API
+impelemented in the python library Flask-RESTful \textbf{???}. GET
+requests are submitted to the database via HTTP, returning json
+formatted allele frequency data and the meta-data associated with each
+population and genetic variant e.g.~latitude, longitude, population
+label, sample size, and frequency scale. Genetic variants can be queried
+by chromosome position, rsid, or randomly. Example HTTP requests and
+json response can be seen below:
+
+\subsection*{Query API by rsid}\label{query-api-by-rsid}
+\addcontentsline{toc}{subsection}{Query API by rsid}
+
+\begin{verbatim}
+http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&rsID=rs1834640
+[
+  {
+    "alleles": ["A", "G"], 
+    "pos": ["-15.310139", "13.443182"], 
+    "pop": "GWD", 
+    "nobs": "226", 
+    "xobs": "17", 
+    "freqscale": 1, 
+    "freq": [0.0752212389381, 0.9247787610619], 
+    "chrom_pos": "15:48392165", 
+    "rawfreq": 0.0752212389381
+  },
+  ...
+]
+\end{verbatim}
+
+\subsection*{Query API by chromosome
+position}\label{query-api-by-chromosome-position}
+\addcontentsline{toc}{subsection}{Query API by chromosome position}
+
+\begin{verbatim}
+http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&chr=14&pos=37690093
+[
+  {
+    "alleles": ["G", "A"], 
+    "pos": ["-15.310139", "13.443182"], 
+    "pop": "GWD", 
+    "nobs": "226", 
+    "xobs": "0", 
+    "freqscale": 0.01, 
+    "freq": [0.0, 1.0], 
+    "chrom_pos": "14:37690093", 
+    "rawfreq": 0.0
+  }
+  ...
+]
+\end{verbatim}
+
+\subsection*{Randomly query API}\label{randomly-query-api}
+\addcontentsline{toc}{subsection}{Randomly query API}
+
+\begin{verbatim}
+http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&random_snp=True
+[
+  {
+    "alleles": ["T", "C"], 
+    "pos": ["-15.310139", "13.443182"], 
+    "pop": "GWD", 
+    "nobs": "226", 
+    "xobs": "0", 
+    "freqscale": 0.01, 
+    "freq": [0.0, 1.0], 
+    "chrom_pos": "5:42452893", 
+    "rawfreq": 0.0
+  }, 
+  ...
+]
+\end{verbatim}
 
@@
 Support for this work was provided by the NIH-BD2K initiative (1U01
-CA198933-0). The authors would also like to thank XX for supportive
-conversations.
-
-\section*{Acknowledgements}\label{acknowledgements-1}
-\addcontentsline{toc}{section}{Acknowledgements}
+CA198933-0). The authors would also like to thank members of the
+Novembre Lab for supportive conversations.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-rosenberg2010genome}{}
+1. Rosenberg NA, Huang L, Jewett EM, Szpiech ZA, Jankovic I, Boehnke M.
+Genome-wide association studies in diverse populations. Nature Reviews
+Genetics. Nature Publishing Group; 2010;11: 356--366.
+
 \hypertarget{ref-novembre2009spatial}{}
-1. Novembre J, Di Rienzo A. Spatial patterns of variation due to natural
+2. Novembre J, Di Rienzo A. Spatial patterns of variation due to natural
 selection in humans. Nature Reviews Genetics. Nature Publishing Group;
@@
 
+\hypertarget{ref-coop2010using}{}
+3. Coop G, Witonsky D, Di Rienzo A, Pritchard JK. Using environmental
+correlations to identify loci underlying local adaptation. Genetics.
+Genetics Soc America; 2010;185: 1411--1423.
+
+\hypertarget{ref-nelson2012abundance}{}
+4. Nelson MR, Wegmann D, Ehm MG, Kessner D, Jean PS, Verzilli C, et al.
+An abundance of rare functional variants in 202 drug target genes
+sequenced in 14,002 people. Science. American Association for the
+Advancement of Science; 2012;337: 100--104.
+
+\hypertarget{ref-mathieson2014demography}{}
+5. Mathieson I, McVean G. Demography and the age of rare variants. PLoS
+Genet. Public Library of Science; 2014;10: e1004528.
+
 \hypertarget{ref-price2006principal}{}
-2. Price AL, Patterson NJ, Plenge RM, Weinblatt ME, Shadick NA, Reich D.
+6. Price AL, Patterson NJ, Plenge RM, Weinblatt ME, Shadick NA, Reich D.
 Principal components analysis corrects for stratification in genome-wide
@@
 \hypertarget{ref-patterson2006population}{}
-3. Patterson N, Price AL, Reich D. Population structure and
+7. Patterson N, Price AL, Reich D. Population structure and
 eigenanalysis. PLoS genet. Public Library of Science; 2006;2: e190.
@@
 \hypertarget{ref-pritchard2000inference}{}
-4. Pritchard JK, Stephens M, Donnelly P. Inference of population
+8. Pritchard JK, Stephens M, Donnelly P. Inference of population
 structure using multilocus genotype data. Genetics. Genetics Soc
@@
 \hypertarget{ref-alexander2009fast}{}
-5. Alexander DH, Novembre J, Lange K. Fast model-based estimation of
+9. Alexander DH, Novembre J, Lange K. Fast model-based estimation of
 ancestry in unrelated individuals. Genome research. Cold Spring Harbor
@@
 \hypertarget{ref-furchedata}{}
-6. Furche T, Gottlob G, Libkin L, Orsi G, Paton NW. Data wrangling for
+10. Furche T, Gottlob G, Libkin L, Orsi G, Paton NW. Data wrangling for
 big data: Challenges and opportunities.
@@
 \hypertarget{ref-rajeevan2011alfred}{}
-7. Rajeevan H, Soundararajan U, Kidd JR, Pakstis AJ, Kidd KK. ALFRED: An
-allele frequency resource for research and teaching. Nucleic acids
+11. Rajeevan H, Soundararajan U, Kidd JR, Pakstis AJ, Kidd KK. ALFRED:
+An allele frequency resource for research and teaching. Nucleic acids
 research. Oxford Univ Press; 2011; gkr924.
@@
 \hypertarget{ref-pickrell2009signals}{}
-8. Pickrell JK, Coop G, Novembre J, Kudaravalli S, Li JZ, Absher D, et
+12. Pickrell JK, Coop G, Novembre J, Kudaravalli S, Li JZ, Absher D, et
 al. Signals of recent positive selection in a worldwide sample of human
@@
 \hypertarget{ref-wessel2013generic}{}
-9. Wessel P, Smith WH, Scharroo R, Luis J, Wobbe F. Generic mapping
+13. Wessel P, Smith WH, Scharroo R, Luis J, Wobbe F. Generic mapping
 tools: Improved version released. Eos, Transactions American Geophysical
@@
 
+\hypertarget{ref-coop2009role}{}
+14. Coop G, Pickrell JK, Novembre J, Kudaravalli S, Li J, Absher D, et
+al. The role of geography in human adaptation. PLoS Genet. Public
+Library of Science; 2009;5: e1000500.
+
 \hypertarget{ref-dudley2013exploring}{}
-10. Dudley JT, Karczewski KJ. Exploring personal genomics. Oxford
+15. Dudley JT, Karczewski KJ. Exploring personal genomics. Oxford
 University Press; 2013.
@@
 \hypertarget{ref-li2008worldwide}{}
-11. Li JZ, Absher DM, Tang H, Southwick AM, Casto AM, Ramachandran S, et
+16. Li JZ, Absher DM, Tang H, Southwick AM, Casto AM, Ramachandran S, et
 al. Worldwide human relationships inferred from genome-wide patterns of
@@
 \hypertarget{ref-10002015global}{}
-12. Consortium 1GP, others. A global reference for human genetic
+17. Consortium 1GP, others. A global reference for human genetic
 variation. Nature. Nature Publishing Group; 2015;526: 68--74.
@@
 \hypertarget{ref-meyer2012high}{}
-13. Meyer M, Kircher M, Gansauge M-T, Li H, Racimo F, Mallick S, et al.
+18. Meyer M, Kircher M, Gansauge M-T, Li H, Racimo F, Mallick S, et al.
 A high-coverage genome sequence from an archaic denisovan individual.
@@
 \hypertarget{ref-lazaridis2014ancient}{}
-14. Lazaridis I, Patterson N, Mittnik A, Renaud G, Mallick S, Kirsanow
+19. Lazaridis I, Patterson N, Mittnik A, Renaud G, Mallick S, Kirsanow
 K, et al. Ancient human genomes suggest three ancestral populations for
@@
 \hypertarget{ref-bostock2011d3}{}
-15. Bostock M, Ogievetsky V, Heer J. D\(^3\) data-driven documents.
+20. Bostock M, Ogievetsky V, Heer J. D\(^3\) data-driven documents.
 Visualization and Computer Graphics, IEEE Transactions on. IEEE;
@@
 \hypertarget{ref-sherry2001dbsnp}{}
-16. Sherry ST, Ward M-H, Kholodov M, Baker J, Phan L, Smigielski EM, et
+21. Sherry ST, Ward M-H, Kholodov M, Baker J, Phan L, Smigielski EM, et
 al. DbSNP: The nCBI database of genetic variation. Nucleic acids
@@
 
-\hypertarget{ref-nelson2012abundance}{}
-17. Nelson MR, Wegmann D, Ehm MG, Kessner D, Jean PS, Verzilli C, et al.
-An abundance of rare functional variants in 202 drug target genes
-sequenced in 14,002 people. Science. American Association for the
-Advancement of Science; 2012;337: 100--104.
-
 \hypertarget{ref-tennessen2012evolution}{}
-18. Tennessen JA, Bigham AW, O'Connor TD, Fu W, Kenny EE, Gravel S, et
+22. Tennessen JA, Bigham AW, O'Connor TD, Fu W, Kenny EE, Gravel S, et
 al. Evolution and functional impact of rare coding variation from deep

### 2016-06-30 paper.Rmd
fixed a few typos

@@
 The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry.
-Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration @nelson2012abundance, @mathieson2014demography.
+Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography.
 Thus by surveying the collection of rare variants that an individual carries provides information about the population origins of that individual's recent ancestors.
@@
 
-## Query API by rsid {-}
+## Query by rsid {-}
 ```
@@
 
-## Query API by chromosome position {-}
+## Query by chromosome position {-}
 
@@
 
-## Randomly query API {-}
+## Random query {-}
 

### 2016-06-30 paper.tex
fixed a few typos

@@
 and have had little time to spread across populations through migration
-{[}4{]}, {[}5{]}. Thus by surveying the collection of rare variants that
-an individual carries provides information about the population origins
-of that individual's recent ancestors. One of us (J.N.) recently took
-advantage of the geographic clustering of an individual's rare variants
-in helping to reaffirm the historically highly contested Native American
-ancestry of the Kennewick Man, an ancient individual from Washington
-State \textbf{???}.
+events {[}4{]}, {[}5{]}. Thus by surveying the collection of rare
+variants that an individual carries provides information about the
+population origins of that individual's recent ancestors. One of us
+(J.N.) recently took advantage of the geographic clustering of an
+individual's rare variants in helping to reaffirm the historically
+highly contested Native American ancestry of the Kennewick Man, an
+ancient individual from Washington State \textbf{???}.
 
@@
 
-\subsection*{Query API by rsid}\label{query-api-by-rsid}
-\addcontentsline{toc}{subsection}{Query API by rsid}
+\subsection*{Query by rsid}\label{query-by-rsid}
+\addcontentsline{toc}{subsection}{Query by rsid}
 
@@
 
-\subsection*{Query API by chromosome
-position}\label{query-api-by-chromosome-position}
-\addcontentsline{toc}{subsection}{Query API by chromosome position}
+\subsection*{Query by chromosome
+position}\label{query-by-chromosome-position}
+\addcontentsline{toc}{subsection}{Query by chromosome position}
 
@@
 
-\subsection*{Randomly query API}\label{randomly-query-api}
-\addcontentsline{toc}{subsection}{Randomly query API}
+\subsection*{Random query}\label{random-query}
+\addcontentsline{toc}{subsection}{Random query}
 

### 2016-07-01 paper.Rmd
filled in more captions

@@
 Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants.
-Through an efficient REST API and dynamic front-end the Geography of Genetic Variants (GGV) browser rapidly provides maps of allele frequencies in populations distributed across the globe.
+Through an efficient REST API and dynamic front-end the [\textit{Geography of Genetic Variants (GGV)}](popgen.uchicago.edu/ggv) browser rapidly provides maps of allele frequencies in populations distributed across the globe.
 
@@
 ```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Schematic diagram of the Geography of Genetic Variants (GGV) web application.
-We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genom Diversity Project (HGDP) and Population Reference Sample (POPRES).
-These allele freqeuncies are stored in a SQL database along side with geographic coordinates associated with each population.
+We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
+These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
 The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.'}
@@
 ```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application.
-Each pie chart represents a population with the blue slice of the pie displaying the freqeuncy of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.'}
+Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.'}
 img <- readPNG("figures/png/figure2.png")
@@
 
-```{r figure3, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Figure caption'}
+```{r figure3, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
+The transparency is scaled in proportion to the number of observed chromosomes for a particular variant in each population.'}
 img <- readPNG("figures/png/figure3.png")
@@
 
-```{r figure4, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Figure caption'}
+```{r figure4, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
+The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.'}
 img <- readPNG("figures/png/figure4.png")
@@
 
-```{r figure5, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Figure caption'}
+```{r figure5, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
+The left map shows the original population locations while the right shows the application of the force directed.'}
 img <- readPNG("figures/png/figure5.png")

### 2016-07-01 paper.tex
filled in more captions

@@
 distribution of genetic variants. Through an efficient REST API and
-dynamic front-end the Geography of Genetic Variants (GGV) browser
-rapidly provides maps of allele frequencies in populations distributed
-across the globe.
+dynamic front-end the
+\href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}}
+browser rapidly provides maps of allele frequencies in populations
+distributed across the globe.
 
@@
 \caption[Schematic diagram of the Geography of Genetic Variants (GGV) web application]{Schematic diagram of the Geography of Genetic Variants (GGV) web application.
-We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genom Diversity Project (HGDP) and Population Reference Sample (POPRES).
-These allele freqeuncies are stored in a SQL database along side with geographic coordinates associated with each population.
+We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
+These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
 The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
@@
 \caption[Example map from the Geography of Genetic Variants web application]{Example map from the Geography of Genetic Variants web application.
-Each pie chart represents a population with the blue slice of the pie displaying the freqeuncy of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:figure2}
+Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:figure2}
 \end{figure}
@@
 
-\caption[Figure caption]{Figure caption}\label{fig:figure3}
+\caption[Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies]{Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
+The transparency is scaled in proportion to the number of observed chromosomes for a particular variant in each population.}\label{fig:figure3}
 \end{figure}
@@
 
-\caption[Figure caption]{Figure caption}\label{fig:figure4}
+\caption[Example map from the Geography of Genetic Variants web application displaying the use of frequency scales for more expressive representations of rare variation on geographic maps]{Example map from the Geography of Genetic Variants web application displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
+The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.}\label{fig:figure4}
 \end{figure}
@@
 
-\caption[Figure caption]{Figure caption}\label{fig:figure5}
+\caption[Example map from the Geography of Genetic Variants web application displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position]{Example map from the Geography of Genetic Variants web application displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
+The left map shows the original population locations while the right shows the application of the force directed.}\label{fig:figure5}
 \end{figure}

### 2016-07-01 paper.Rmd
upated readme and fixed typo

@@
 ```{r figure3, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
-The transparency is scaled in proportion to the number of observed chromosomes for a particular variant in each population.'}
+The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.'}
 img <- readPNG("figures/png/figure3.png")

### 2016-07-01 paper.Rmd
fixing figure formatting

@@
 csl: plos.csl
+header-includes:
+   - \usepackage{amsmath}
+   - \usepackage{amsthm}
+   - \usepackage{float}
 bibliography: library.bib
@@
 Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants.
-Through an efficient REST API and dynamic front-end the [\textit{Geography of Genetic Variants (GGV)}](popgen.uchicago.edu/ggv) browser rapidly provides maps of allele frequencies in populations distributed across the globe.
+Through an efficient SQL database, REST API and dynamic front-end the [\textit{Geography of Genetic Variants (GGV)}](popgen.uchicago.edu/ggv) browser rapidly provides maps of allele frequencies in populations distributed across the globe.
 
@@
 
-# Fundamental Approach {-}
-
-The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of D3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
-
-```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Schematic diagram of the Geography of Genetic Variants (GGV) web application.
+```{r figure1, echo=FALSE, message=FALSE, fig.height=4, fig.width=4, fig.align='center', fig.cap='Schematic diagram of the Geography of Genetic Variants (GGV) browser.
 We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
@@
 ```
+
+# Fundamental Approach {-}
+
+The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of D3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
 
@@
 
-```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application.
+```{r figure2, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants browser.
 Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.'}
@@
 
-```{r figure3, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
+```{r figure3, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
 The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.'}
@@
 
+```{r figure4, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
+The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.'}
+img <- readPNG("figures/png/figure4.png")
+grid.raster(img)
+```
+
 In current visualization schemes, such as the HGDP Selection Browser, rare variants would be represented as narrow slivers in a pie chart, nearly invisible to the naked eye.
@@
 Much like scientific notation, this allows a wide range of frequencies to be displayed].
-
-```{r figure4, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
-The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.'}
-img <- readPNG("figures/png/figure4.png")
-grid.raster(img)
-```
 
@@
 
-```{r figure5, echo=FALSE, message=FALSE, fig.height=6, fig.width=6, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
-The left map shows the original population locations while the right shows the application of the force directed.'}
+```{r figure5, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
+The left map shows the original population locations while the right shows the application of the force directed layout.'}
 img <- readPNG("figures/png/figure5.png")

### 2016-07-01 paper.tex
upated readme and fixed typo
fixing figure formatting

@@
 an interactive visualization tool for illuminating the geographic
-distribution of genetic variants. Through an efficient REST API and
-dynamic front-end the
+distribution of genetic variants. Through an efficient SQL database,
+REST API and dynamic front-end the
 \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}}
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/figure1-1}
+
+}
+
+\caption[Schematic diagram of the Geography of Genetic Variants (GGV) browser]{Schematic diagram of the Geography of Genetic Variants (GGV) browser.
+We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
+These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
+The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
+\end{figure}
+\end{CodeChunk}
+
 \section*{Fundamental Approach}\label{fundamental-approach}
@@
 variant.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/figure1-1}
-
-}
-
-\caption[Schematic diagram of the Geography of Genetic Variants (GGV) web application]{Schematic diagram of the Geography of Genetic Variants (GGV) web application.
-We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
-These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
-\end{figure}
-\end{CodeChunk}
 
@@
 
-\caption[Example map from the Geography of Genetic Variants web application]{Example map from the Geography of Genetic Variants web application.
+\caption[Example map from the Geography of Genetic Variants browser]{Example map from the Geography of Genetic Variants browser.
 Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:figure2}
@@
 
-\caption[Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies]{Example map from the Geography of Genetic Variants web application displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
-The transparency is scaled in proportion to the number of observed chromosomes for a particular variant in each population.}\label{fig:figure3}
+\caption[Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies]{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
+The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.}\label{fig:figure3}
 \end{figure}
@@
 
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/figure4-1}
+
+}
+
+\caption[Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps]{Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
+The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.}\label{fig:figure4}
+\end{figure}
+\end{CodeChunk}
+
 In current visualization schemes, such as the HGDP Selection Browser,
@@
 notation, this allows a wide range of frequencies to be displayed{]}.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/figure4-1}
-
-}
-
-\caption[Example map from the Geography of Genetic Variants web application displaying the use of frequency scales for more expressive representations of rare variation on geographic maps]{Example map from the Geography of Genetic Variants web application displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
-The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.}\label{fig:figure4}
-\end{figure}
-\end{CodeChunk}
 
@@
 
-\caption[Example map from the Geography of Genetic Variants web application displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position]{Example map from the Geography of Genetic Variants web application displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
-The left map shows the original population locations while the right shows the application of the force directed.}\label{fig:figure5}
+\caption[Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position]{Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
+The left map shows the original population locations while the right shows the application of the force directed layout.}\label{fig:figure5}
 \end{figure}

### 2016-07-01 paper.Rmd
reformatted figures bib and fixed some typos

@@
 Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography.
-Thus by surveying the collection of rare variants that an individual carries provides information about the population origins of that individual's recent ancestors.
-One of us (J.N.) recently took advantage of the geographic clustering of an individual's rare variants in helping to reaffirm the historically highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.
+By surveying the collection of rare variants that an individual carries one can gain insight into the geographic origins of that individual's recent ancestors.
+One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically-highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.
 
 Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets.
-For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide vairants (SNVs).
+For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
 One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast.
@@
 Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.
-One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a sbustantial learning curve.
+One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a substantial learning curve.
 The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @pickrell2009signals, @coop2009role, books e.g.
-@dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
+@dudley2013exploring, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
 
@@
 # Underlying Frequency API {-}
-To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API impelemented in the python library Flask-RESTful @citeXXX.
+To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API implemented in the python library Flask-RESTful @citeXXX.
 GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g.
@@
 While doing so is generally immensely helpful, it has inherent complexity.
-For example,  pracitioners must make choices regarding representing where an individual was sampled for the study (e.g.
+For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g.
 the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
@@
 # References {-}
+
+<!-- Has to be set for some wierd indentation issue -->
+\setlength\parindent{0pt}

### 2016-07-01 paper.tex
reformatted figures bib and fixed some typos

@@
 and have had little time to spread across populations through migration
-events {[}4{]}, {[}5{]}. Thus by surveying the collection of rare
-variants that an individual carries provides information about the
-population origins of that individual's recent ancestors. One of us
-(J.N.) recently took advantage of the geographic clustering of an
-individual's rare variants in helping to reaffirm the historically
-highly contested Native American ancestry of the Kennewick Man, an
-ancient individual from Washington State \textbf{???}.
+events {[}4{]}, {[}5{]}. By surveying the collection of rare variants
+that an individual carries one can gain insight into the geographic
+origins of that individual's recent ancestors. One of us (J.N.) recently
+took advantage of the geographic localization of an individual's rare
+variants in helping to reaffirm the historically-highly contested Native
+American ancestry of the Kennewick Man, an ancient individual from
+Washington State \textbf{???}.
 
@@
 of thousands of individuals measured at hundreds of thousands or even
-millions of single nucleotide vairants (SNVs). One common approach to
+millions of single nucleotide variants (SNVs). One common approach to
 visualizing such high-dimensional data is to compress the SNV dimensions
@@
 The Generic Mapping Tools (GMT) {[}13{]}, a powerful system of
-geographic plotting scripts with a sbustantial learning curve. The plots
+geographic plotting scripts with a substantial learning curve. The plots
 from the HGDP Selection Browser have proved useful, and have appeared in
 research articles e.g {[}12{]}, {[}14{]}, books e.g. {[}15{]}, and have
-been ported and made avialable on the UCSC Genome Browser (available
+been ported and made available on the UCSC Genome Browser (available
 under the HGDP Allele Freq track of the browser).
@@
 database with the front-end D3.js visualization we use a REST API
-impelemented in the python library Flask-RESTful \textbf{???}. GET
+implemented in the python library Flask-RESTful \textbf{???}. GET
 requests are submitted to the database via HTTP, returning json
@@
 geographic location. While doing so is generally immensely helpful, it
-has inherent complexity. For example, pracitioners must make choices
+has inherent complexity. For example, practitioners must make choices
 regarding representing where an individual was sampled for the study
@@
 \addcontentsline{toc}{section}{References}
+
+\setlength\parindent{0pt}
 

### 2016-07-01 paper.Rmd
updated typos

@@
 Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography.
-By surveying the collection of rare variants that an individual carries one can gain insight into the geographic origins of that individual's recent ancestors.
-One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically-highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.
+Surveying the collection of rare variants that an individual carries  can provide insight into the geographic origins of that individual's recent ancestors.
+One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.
 
@@
 For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
-One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast.
+One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, or a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast.
 While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail.
@@
 In addition, novel approaches for data visualization have become more widely available.
-In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.
+In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing @bostock2011d3.
 

### 2016-07-01 paper.tex
updated typos

@@
 and have had little time to spread across populations through migration
-events {[}4{]}, {[}5{]}. By surveying the collection of rare variants
-that an individual carries one can gain insight into the geographic
-origins of that individual's recent ancestors. One of us (J.N.) recently
-took advantage of the geographic localization of an individual's rare
-variants in helping to reaffirm the historically-highly contested Native
+events {[}4{]}, {[}5{]}. Surveying the collection of rare variants that
+an individual carries can provide insight into the geographic origins of
+that individual's recent ancestors. One of us (J.N.) recently took
+advantage of the geographic localization of an individual's rare
+variants in helping to reaffirm the historically contested Native
 American ancestry of the Kennewick Man, an ancient individual from
@@
 down to a small number of latent factors, using a method such as
-principal components analysis {[}6{]}, {[}7{]}, a model-based clustering
-method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While these
-approaches are extremely valuable, researchers can use them to often
-without inspecting the underlying variant data in more detail. A natural
-approach to gaining more insight to the overall structure of a
+principal components analysis {[}6{]}, {[}7{]}, or a model-based
+clustering method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While
+these approaches are extremely valuable, researchers can use them to
+often without inspecting the underlying variant data in more detail. A
+natural approach to gaining more insight to the overall structure of a
 population genetic dataset is to visually inspect what geographic
@@
 software development in modern web-browsers, a large open-source
-development community, and ease of share-ability {[}20{]}.
+development community, and ease of sharing {[}20{]}.
 

### 2016-07-01 paper.Rmd
fixing more typos

@@
 Reference datasets for population genetic variation has greatly expanded since the release of the HGDP Illumina 650Y dataset @li2008worldwide that formed the basis of the HGDP Selection Browser maps.
-The most notable advance is the publication of the 1000 Genomes phase 3 data @10002015global though additional datasets are coming online @meyer2012high, @lazaridis2014ancient.
+The most notable advance is the publication of the 1000 Genomes Phase 3 data @10002015global though additional datasets are coming online @meyer2012high, @lazaridis2014ancient.
 
@@
 These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.'}
+The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.'}
 img <- readPNG("figures/png/figure1.png")
@@
 
-```{r figure4, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
+```{r figure4, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
 The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.'}
@@
 
-```{r figure5, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
+```{r figure5, echo=FALSE, message=FALSE, fig.height=6, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
 The left map shows the original population locations while the right shows the application of the force directed layout.'}

### 2016-07-01 paper.tex
fixing more typos

@@
 the basis of the HGDP Selection Browser maps. The most notable advance
-is the publication of the 1000 Genomes phase 3 data {[}17{]} though
+is the publication of the 1000 Genomes Phase 3 data {[}17{]} though
 additional datasets are coming online {[}18{]}, {[}19{]}.

### 2016-07-01 paper.tex
updated

@@
 These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
+The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
 \end{figure}

### 2016-07-05 paper.Rmd
Some edits / comments

@@
 # Abstract {-}
-One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its frequency, is its geographic distribution.
-The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it spread within, and in turn can help one learn about historical patterns of migration, genetic drift, and natural selection.
-Collectively the geographic distribution of genetic variants can help to explain how populations have been related through time (e.g.
-levels of gene flow and divergence).
-For these reasons, visual inspection of geographic maps for genetic variants is common practice in genetic studies.
+
+One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its global frequency, is its geographic distribution.
+The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it has spread to, and in turn how migration, genetic drift, and natural selection have acted on the variant.
+Collectively the
+geographic distribution of many genetic variants can inform on how populations have been related through time (e.g.~levels of gene flow and divergence).
+The distribution of a genetic variant can also be informative background knowledge for medical/clinical geneticists.
+As a result, visual
+inspection of geographic maps for genetic variants is common practice in genetic studies.
 Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants.
@@
 # Introduction {-}
-Genetics researchers often face the following problem: the researcher has identified one or many genetic variants of interest (e.g.
+Genetics researchers often face the problem that they have identified one or many genetic variants of interest (e.g.
 from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant.
@@
 
-The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry.
-Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography.
-Surveying the collection of rare variants that an individual carries  can provide insight into the geographic origins of that individual's recent ancestors.
-One of us (J.N.) recently took advantage of the geographic localization of an individual's rare variants in helping to reaffirm the historically contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.
+The geography of genetic variants is also the basis for many supervised methods for individual-level ancestry inference.
+The constellation of common single-nucleotide polymorphism (SNP) alleles carried by an individual can provide some insight to the geographic origins of their recent ancestors.
+More recently, the increasing availability of data on rare variants is opening up new opportunities for ancestry inference.
+Rare variants are particularly informative for ancestry as they tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events @nelson2012abundance, @mathieson2014demography.
+By assessing the geographic distribution of rare variants carried by an individual one can gain information about their ancestry (e.g.
+BrownPasniuc2014PloSCompBio, @KennewickManReport).
+In these applications, geographic maps of the genetic variants an individual carries can be useful and informative.
+[xxJN: I have mixed feelings about this paragraph - while GGV/GGVapi have relevance for ancestry questions, it's not set up for users to use it that way; I think it could be omitted or possibly moved to discussion as a future direction) ]
 
@@
 One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis @price2006principal, @patterson2006population, or a model-based clustering method such as STRUCTURE @pritchard2000inference or ADMIXTURE @alexander2009fast.
-While these approaches are extremely valuable, researchers can use them to often without inspecting the underlying variant data in more detail.
+While these approaches are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
 A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.
 
-Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods @furchedata and geographic plotting techniques that are unfamiliar to most.
+Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods @furchedata and map making techniques that are unfamiliar to most.
 Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.
@@
 Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.
-One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a substantial learning curve.
+One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts for making static plots.
 The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @pickrell2009signals, @coop2009role, books e.g.
@@
 
-Reference datasets for population genetic variation has greatly expanded since the release of the HGDP Illumina 650Y dataset @li2008worldwide that formed the basis of the HGDP Selection Browser maps.
-The most notable advance is the publication of the 1000 Genomes Phase 3 data @10002015global though additional datasets are coming online @meyer2012high, @lazaridis2014ancient.
+Reference datasets for population genetic variation have greatly expanded since the release of the HGDP Illumina 650Y dataset @li2008worldwide that formed the basis of the HGDP Selection Browser maps.
+The most notable advance is the publication of the 1000 Genomes Phase 3 data @10002015global though additional datasets are continually coming online @meyer2012high, @lazaridis2014ancient.
+In addition, novel approaches for data visualization have become more widely available.
+In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer useful methods for interactivity, the advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing @bostock2011d3.
 
-In addition, novel approaches for data visualization have become more widely available.
-In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing @bostock2011d3.
+Taking advantage of these recent advances, we aim to address the significant visualization challenges that are inherit in the production of geographic allele frequency maps, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
 
-Taking advantage of these recent advances, we aim to address significant visualization challenges that are inherit in the production of geographic allele frequency maps, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
-
-```{r figure1, echo=FALSE, message=FALSE, fig.height=4, fig.width=4, fig.align='center', fig.cap='Schematic diagram of the Geography of Genetic Variants (GGV) browser.
+```{r figure1, echo=FALSE, message=FALSE, fig.height=4, fig.width=4, fig.align='center', fig.cap=' [xxJN= This figure can be made much smaller - the layout would be more compact either be left-to-right, or up-and-down, as opposed to diaganol or even better a triangle so that the whole image is square and would take up approximately a column width of a final paper.
+Also decrease font and icon size.] Schematic diagram of the Geography of Genetic Variants (GGV) browser.
 We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
@@
 
-The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of D3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
+The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics (SVG [xxJN: double check correct expansion of SVG acronym]) and mapping utilities of D3.js @bostock2011d3 to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
 
-In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project, Human Genome Diversity project and POPRES project, we have developed an SQL database and REST API for querying allele frequencies by chromosome and position or reference SNP identifier [Figure 1].
+In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project @10002015global, Human Genome Diversity project @li2008worldwide, we have developed an SQL database and REST API for querying allele frequencies by chromosome and position or by reference SNP identifier @sherry2001dbsnp [Figure 1].
 
-Users can query single variants by chromosome and position identifiers, by rsid @sherry2001dbsnp, or users can simply choose a random variant from within a dataset.
+The front-end of the project is a browser-based visualization application that displays the geographci map of a particular variant and provides legends for the map and various boxes to allow users to query different datasets or choose visualization options.
+Here, users can query single variants by chromosome and position identifiers, by rsid @sherry2001dbsnp, or users can simply choose a random variant from within a dataset [xxJN: note level of detail in this setence is somewhat redundant with one in abvoe paragaph about API; and can API provide random variants as well?].
 While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
@@
 
-```{r figure2, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants browser.
+```{r figure2, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='[xxJN: I think first figure would be stronger as a screenshot of the whole display field for the browser app; what do you think?] Example map from the Geography of Genetic Variants browser using the @li2008worldwide data.
 Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.'}
@@
 After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies.
-Pie charts are displayed as points positioned at the population's region of origin where the projection / map-view is chosen based off of the geographic proximity of populations present in a given dataset [Figure 2].
+Pie charts are displayed as points positioned at the population's region of origin and the projection / map-view is chosen based off of the geographic configuration of populations in a given dataset [Figure 2].
 
-# Representing variable certainty in frequency data {-}
+# Representing uncertainty in frequency data {-}
 One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
 For some locations, sample sizes are small, and the reported allele frequency may be quite far from the true population frequency due to sampling error.
-To address this issue, we use varying transparency in a population's pie chart: sample allele frequencies with higher levels of sampling error will be made more transparent, and hence less visible on the map [Figure 3].
-
+To address this issue, we use varying transparency in a population's pie chart: sampled populations with higher levels of sampling error (i.e.
+n<30) are made more transparent, and hence less visible on the map [Figure 3].
 ```{r figure3, echo=FALSE, message=FALSE, fig.height=4, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
-The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.'}
+The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.
+The frequency data and population identifiers are from @novembreetal2008.'}
 img <- readPNG("figures/png/figure3.png")
@@
 An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset.
-This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays, biased towards variants that are common in human populations (5-50% in minor allele frequency).
-With the advent of next generation sequencing technologies and large samples of thousands of individuals, it is common for datasets to contain rare variants @10002015global, @nelson2012abundance, @tennessen2012evolution.
+This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays, which biased towards variants that are common in human populations (5-50% in minor allele frequency).
+With the advent of next generation sequencing technologies, novel array designs focusing on rarer variants, and large samples of thousands of individuals, it is now common for datasets to contain rare variants @10002015global, @nelson2012abundance, @tennessen2012evolution.
+In current visualization schemes, such as the HGDP Selection Browser, rare variants would be represented as narrow slivers in a pie chart, nearly invisible to the naked eye.
 
 ```{r figure4, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
-The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.'}
+The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.
+The data are from @10002015global.'}
 img <- readPNG("figures/png/figure4.png")
@@
 
-In current visualization schemes, such as the HGDP Selection Browser, rare variants would be represented as narrow slivers in a pie chart, nearly invisible to the naked eye.
 To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
-Specifically, we will use a “frequency scale” that is indicated below the map, and redundantly using color, that will indicate a constant scale for the frequencies displayed [Figure 4].
-Much like scientific notation, this allows a wide range of frequencies to be displayed].
+Specifically, we  use a “frequency scale” that is indicated in a legend below the map, and redundantly by using color, that indicates the scale for the frequencies displayed [Figure 4].
+Much like scientific notation, this allows a wide range of frequencies to be displayed.
+[xxJN: I'd like us to use the example maps, plus a figure that shows a table of the colors being used and how they translate.
+I have a figure I've used in talks that should work]
 
-# Interface features {-}
+# Additional features of the interface {-}
 In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
-To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins.
-Lines anchoring the points visibly to their original sampling locations will be used to make sure true sampling locations are indicated [Figure 5].
+To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure 5].
+Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
 

### 2016-07-08 paper.Rmd
More edits

@@
 
-# Underlying Frequency API {-}
-To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API implemented in the python library Flask-RESTful @citeXXX.
-GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g.
+# Access to the underlying frequency data {-}
+To provide an interface to the population minor allele frequency data stored in the SQL database we use a REST API implemented in the python library Flask-RESTful @citeXXX.
+The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
+For the front-end, GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g.
 latitude, longitude, population label, sample size, and frequency scale.
 Genetic variants can be queried by chromosome position, rsid, or randomly.
-Example HTTP requests and json response can be seen below:
+Example HTTP requests and json response can be seen in Examples 1,2,3 (below).
 
-## Query by rsid {-}
+# Caveats {-}
+A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.
+While doing so is generally immensely helpful, it has inherent complexity and limitations.
+For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g.
+the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
+as determined in practice by the birthplaces of recent ancestors, such as grandparents).
+We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that is based on those used by the initial analysts of the data.
+A future feature will allow alternative location schema to be used for the populations in a dataset.
+
+# Conclusion {-}
+By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
+We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
+
+# Acknowledgements {-}
+Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank members of the Novembre Lab for supportive conversations.
+
+# References {-}
+
+<!-- Has to be set for some wierd indentation issue -->
+\setlength\parindent{0pt}
+
+## Example 1: Query by rsid {-}
 ```
@@
 
-## Query by chromosome position {-}
+## Example 2: Query by chromosome position {-}
 
@@
 
-## Random query {-}
+## Example 3: Random query {-}
 
@@
 ```
-
-# Caveats {-}
-A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.
-While doing so is generally immensely helpful, it has inherent complexity.
-For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g.
-the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
-as determined in practice by the birthplaces of recent ancestors, such as grandparents).
-We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that aligns with those used by the initial analysts of the data.
-A future feature will be to allow alternative location schema to be used for the populations in a dataset.
-
-# Conclusion {-}
-By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
-We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
-
-# Acknowledgements {-}
-Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank members of the Novembre Lab for supportive conversations.
-
-# References {-}
-
-<!-- Has to be set for some wierd indentation issue -->
-\setlength\parindent{0pt}

### 2016-07-16 paper.tex
reformatted paper, changed figures to pdf, working through JN commnts. started freq table

--- 2016-07-01 paper.tex
@@
-% Template for PLoS
-
-\documentclass[10pt]{article}
-
-% amsmath package, useful for mathematical formulas
-\usepackage{amsmath}
-% amssymb package, useful for mathematical symbols
-\usepackage{amssymb}
-
-% hyperref package, useful for hyperlinks
+\documentclass{article}
+\usepackage{fullpage}
+\usepackage{amsmath,amssymb}
+\usepackage{natbib}
+\usepackage{graphicx}
 \usepackage{hyperref}
-
-% graphicx package, useful for including eps and pdf graphics
-% include graphics with the command \includegraphics
-\usepackage{graphicx}
-
-% Sweave(-like)
-\usepackage{fancyvrb}
-\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
-\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
-\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
-\newenvironment{Schunk}{}{}
-\DefineVerbatimEnvironment{Code}{Verbatim}{}
-\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
-\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
-\newenvironment{CodeChunk}{}{}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
 \usepackage{color}
-
-% Use doublespacing - comment out for single spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\topmargin 0.0cm
-\oddsidemargin 0.5cm
-\evensidemargin 0.5cm
-\textwidth 16cm 
-\textheight 21cm
-
-% Bold the 'Figure #' in the caption and separate it with a period
-% Captions will be left justified
-\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
-
-% Use the PLoS provided bibtex style
-\bibliographystyle{plos}
-
-% Remove brackets from numbering in List of References
+\usepackage{lineno}
+\usepackage{authblk}
+\usepackage{array}
+\usepackage[labelfont=bf]{caption}
 \makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
+\renewcommand*{\@fnsymbol}[1]{\ensuremath{\ifcase#1\or *\or**\or\dagger\or \ddagger\or
+   \mathsection\or \mathparagraph\or \|\or **\or \dagger\dagger
+   \or \ddagger\ddagger \else\@ctrerr\fi}}
 \makeatother
 
-% Leave date blank
+\linenumbers
+\captionsetup{labelfont=bf}
+\frenchspacing
+\title{Visualizing the Geography of Genetic Variants}
+\author[1]{Joseph H. Marcus$^{*}$}
+\author[1]{John Novembre\thanks{Address correspondence to JM (\texttt{josephhmarcus@gmail.com}) or JN (\texttt{jnovembre@uchicago.edu}).}}
+\affil[1]{\small{Department of Human Genetics, University of Chicago, Chicago, IL,USA}}
+
 \date{}
+\begin{document}
+\maketitle
 
-\pagestyle{myheadings}
-%% ** EDIT HERE **
+\begin{abstract}
+One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its global frequency, is its geographic distribution.
+The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it has spread to, and in turn how migration, genetic drift, and natural selection have acted on the variant.
+Collectively the geographic distribution of many genetic variants can inform on how populations have been related through time (e.g.~levels of gene flow and divergence).
+The distribution of a genetic variant can also be informative background knowledge for medical/clinical geneticists.
+As a result, visual inspection of geographic maps for genetic variants is common practice in genetic studies.
+Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants.
+Through an efficient SQL database, REST API and dynamic front-end the \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}} browser rapidly provides maps of allele frequencies in populations distributed across the globe.
+\end{abstract}
+\newpage
 
-%% ** EDIT HERE **
-%% PLEASE INCLUDE ALL MACROS BELOW
+%%%%%%%%%%%%%%%%%%%%%%%
 
-%% END MACROS SECTION
+\section*{Introduction}
 
-\begin{document}
+Genetics researchers often face the problem that they have identified one or many genetic variants of interest (e.g.
+from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant.
+For example, the researcher may hope to address: 1) implications for genomic medicine (e.g.
+is a risk allele geographically localized to a certain patient population?); 2) design of follow-up studies (e.g.
+what population should be studied to observe variant carriers?); or 3) the biology of the variant in question (e.g.
+does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?)  \cite{rosenberg2010genome, novembre2009spatial, coop2010using}.
+A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
 
-% Title must be 150 characters or less
-\begin{flushleft}
-{\Large
-\textbf{Visualizing the geography of genetic variants}
-}
-% Insert Author names, affiliations and corresponding author email.
-\\
-  Joseph H. Marcus\textsuperscript{1*}, 
-  John Novembre\textsuperscript{1*}\\
-\bf{1} Department of Human Genetics, University of Chicago,  Chicago,  IL,  USA
-\\
+Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets.
+For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
+One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \cite{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \cite{pritchard2000inference} or ADMIXTURE \cite{alexander2009fast}.
+While these approaches are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
+A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.
 
-\textasteriskcentered{} E-mail:   \href{mailto:jhmarcus@uchicago.edu}{\nolinkurl{jhmarcus@uchicago.edu}}
-  \href{mailto:jnovembre@uchicago.edu}{\nolinkurl{jnovembre@uchicago.edu}}
+Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods \cite{furchedata} and map making techniques that are unfamiliar to most.
+Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.
 
-\end{flushleft}
+Our work is inspired by past tools such as the ALFRED database \cite{rajeevan2011alfred} and the maps available on the HGDP Selection browser \cite{pickrell2009signals}.
+One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \cite{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
+The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g \cite{pickrell2009signals}, \cite{coop2009role}, books e.g.
+\cite{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
+Reference datasets for population genetic variation have greatly expanded since the release of the HGDP Illumina 650Y dataset \cite{li2008worldwide} that formed the basis of the HGDP Selection Browser maps.
+The most notable advance is the publication of the 1000 Genomes Phase 3 data \cite{10002015global} though additional datasets are continually coming online \cite{meyer2012high, lazaridis2014ancient}.
+In addition, novel approaches for data visualization have become more widely available.
+In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer useful methods for interactivity, the advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing \cite{bostock2011d3}.
 
-\section*{Abstract}\label{abstract}
-\addcontentsline{toc}{section}{Abstract}
-
-One of the key characteristics of any genetic variant, beyond its
-potential phenotypic effects or its frequency, is its geographic
-distribution. The geographic distribution of a genetic variant can shed
-light on where the variant first arose, in what populations it spread
-within, and in turn can help one learn about historical patterns of
-migration, genetic drift, and natural selection. Collectively the
-geographic distribution of genetic variants can help to explain how
-populations have been related through time (e.g.~levels of gene flow and
-divergence). For these reasons, visual inspection of geographic maps for
-genetic variants is common practice in genetic studies. Here we develop
-an interactive visualization tool for illuminating the geographic
-distribution of genetic variants. Through an efficient SQL database,
-REST API and dynamic front-end the
-\href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}}
-browser rapidly provides maps of allele frequencies in populations
-distributed across the globe.
-
-\section*{Introduction}\label{introduction}
-\addcontentsline{toc}{section}{Introduction}
-
-Genetics researchers often face the following problem: the researcher
-has identified one or many genetic variants of interest (e.g.~from a
-genome-wide association, eQTL, or pharmacogenomic study) and would like
-to know the geographic distribution of the variant. For example, the
-researcher may hope to address: 1) implications for genomic medicine
-(e.g.~is a risk allele geographically localized to a certain patient
-population?); 2) design of follow-up studies (e.g.~what population
-should be studied to observe variant carriers?); or 3) the biology of
-the variant in question (e.g.~does the variant correlate with a known
-environmental factor in a manner suggestive of some geographically
-localized selection pressure?) {[}1{]}, {[}2{]}, {[}3{]}. A simple
-geographic map of the distribution of a genetic variant can be
-incredibly insightful for these questions.
-
-The geographic distribution of rare genetic variants (minor allele
-frequency \(\leq .05\)) can be particularly informative in determining
-an individual's ancestry. Rare variants tend to be localized in small
-geographic regions because they have arisen recently in human history
-and have had little time to spread across populations through migration
-events {[}4{]}, {[}5{]}. Surveying the collection of rare variants that
-an individual carries can provide insight into the geographic origins of
-that individual's recent ancestors. One of us (J.N.) recently took
-advantage of the geographic localization of an individual's rare
-variants in helping to reaffirm the historically contested Native
-American ancestry of the Kennewick Man, an ancient individual from
-Washington State \textbf{???}.
-
-Contemporary population genetics researchers are also faced with the
-challenge of large, high dimensional datasets. For example, is not
-uncommon for a researcher in human genetics to have a dataset comprised
-of thousands of individuals measured at hundreds of thousands or even
-millions of single nucleotide variants (SNVs). One common approach to
-visualizing such high-dimensional data is to compress the SNV dimensions
-down to a small number of latent factors, using a method such as
-principal components analysis {[}6{]}, {[}7{]}, or a model-based
-clustering method such as STRUCTURE {[}8{]} or ADMIXTURE {[}9{]}. While
-these approaches are extremely valuable, researchers can use them to
-often without inspecting the underlying variant data in more detail. A
-natural approach to gaining more insight to the overall structure of a
-population genetic dataset is to visually inspect what geographic
-patterns arise in allele frequency maps.
-
-Unfortunately, generating geographic allele frequency maps is
-time-consuming for the average researcher as it requires a combination
-of data wrangling methods {[}10{]} and geographic plotting techniques
-that are unfamiliar to most. Our aim here is to produce a tailored
-system for rapidly constructing informative geographic maps of allele
-frequency variation.
-
-Our work is inspired by past tools such as the ALFRED database {[}11{]}
-and the maps available on the HGDP Selection browser {[}12{]}. One of us
-(J.N.) developed the scripts for the HGDP Selection Browser maps using
-The Generic Mapping Tools (GMT) {[}13{]}, a powerful system of
-geographic plotting scripts with a substantial learning curve. The plots
-from the HGDP Selection Browser have proved useful, and have appeared in
-research articles e.g {[}12{]}, {[}14{]}, books e.g. {[}15{]}, and have
-been ported and made available on the UCSC Genome Browser (available
-under the HGDP Allele Freq track of the browser).
-
-Reference datasets for population genetic variation has greatly expanded
-since the release of the HGDP Illumina 650Y dataset {[}16{]} that formed
-the basis of the HGDP Selection Browser maps. The most notable advance
-is the publication of the 1000 Genomes Phase 3 data {[}17{]} though
-additional datasets are coming online {[}18{]}, {[}19{]}.
-
-In addition, novel approaches for data visualization have become more
-widely available. In particular, web-based visualization tools, such as
-Data Driven Documents (D3.js), offer powerful approaches to realize
-these aims through useful methods for interactivity, advantages of
-software development in modern web-browsers, a large open-source
-development community, and ease of sharing {[}20{]}.
-
-Taking advantage of these recent advances, we aim to address significant
-visualization challenges that are inherit in the production of
-geographic allele frequency maps, including dynamic interaction, display
-of rare genetic variation, and representation of uncertainty in
-estimated allele frequencies due to variable sample sizes.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/figure1-1}
-
-}
-
-\caption[Schematic diagram of the Geography of Genetic Variants (GGV) browser]{Schematic diagram of the Geography of Genetic Variants (GGV) browser.
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=.4\textwidth]{figures/pdf/figure1.pdf}
+ \caption{Schematic diagram of the Geography of Genetic Variants (GGV) browser.
 We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
 These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}\label{fig:figure1}
+The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}
 \end{figure}
-\end{CodeChunk}
 
-\section*{Fundamental Approach}\label{fundamental-approach}
-\addcontentsline{toc}{section}{Fundamental Approach}
+Taking advantage of these recent advances, we aim to address the significant visualization challenges that are inherit in the production of geographic allele frequency maps, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
 
-The Geography of Genetic Variants browser (GGV) uses the SVG and mapping
-utilities of D3.js to generate interactive frequency maps, allowing for
-quick and dynamic displays of the geographic distribution of a genetic
-variant.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-In order to allow for easy access to large commonly used public genomic
-datasets, such as the 1000 Genomes project, Human Genome Diversity
-project and POPRES project, we have developed an SQL database and REST
-API for querying allele frequencies by chromosome and position or
-reference SNP identifier {[}Figure 1{]}.
+\section*{Fundamental Approach}
 
-Users can query single variants by chromosome and position identifiers,
-by rsid {[}21{]}, or users can simply choose a random variant from
-within a dataset. While many applications require inspection of the
-distribution of a specific variant, from our experience, it can be very
-helpful to quickly view the geographic distribution of several randomly
-chosen variants to quickly gain a sense of structure in a dataset. For
-instance, in data with deep population subdivision, it is obvious in the
-consistent patterns of differentiation observed across markers. We also
-expect this will be useful in teaching contexts -- as it provides a
-highly visual way for learners to understand human genetic variation.
+The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \cite{bostock2011d3} to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
 
-\begin{CodeChunk}
-\begin{figure}
+In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \cite{10002015global}, Human Genome Diversity project \cite{li2008worldwide}, we have developed an SQL database and REST API for querying allele frequencies by chromosome and position, by reference SNP identifier \cite{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1].
 
-{\centering \includegraphics{paper_files/figure-latex/figure2-1}
+The front-end of the project is a browser-based visualization application that displays the geographic map of a particular variant and provides legends for the map and various boxes to allow users to query different datasets or choose visualization options.
+While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
+For instance, in data with deep population subdivision, it is obvious in the consistent patterns of differentiation observed across markers.
+We also expect this will be useful in teaching contexts – as it provides a highly visual way for learners to understand human genetic variation.
 
-}
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=.55\textwidth]{figures/png/figure2.png}
+ \caption{Example screenshot from the Geography of Genetic Variants browser using the \cite{10002015global} data.
+Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}
+\end{figure}
 
-\caption[Example map from the Geography of Genetic Variants browser]{Example map from the Geography of Genetic Variants browser.
-Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:figure2}
+After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies.
+Pie charts are displayed as points positioned at the population's region of origin and the projection / map-view is chosen based off of the geographic configuration of populations in a given dataset [Figure 2].
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Representing uncertainty in frequency data}
+
+One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
+For some locations, sample sizes are small, and the reported allele frequency may be quite far from the true population frequency due to sampling error.
+To address this issue, we use varying transparency in a population's pie chart: sampled populations with higher levels of sampling error (i.e.
+$n < 30$) are made more transparent, and hence less visible on the map [Figure 3].
+
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=.65\textwidth]{figures/pdf/figure3.pdf}
+ \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
+The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.
+The frequency data and population identifiers are from \cite{novembre2008genes}.}
 \end{figure}
-\end{CodeChunk}
 
-After a query, the GGV displays the frequency of a variant in a given
-population as a pie chart where each slice represents minor and major
-allele frequencies. Pie charts are displayed as points positioned at the
-population's region of origin where the projection / map-view is chosen
-based off of the geographic proximity of populations present in a given
-dataset {[}Figure 2{]}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section*{Representing variable certainty in frequency
-data}\label{representing-variable-certainty-in-frequency-data}
-\addcontentsline{toc}{section}{Representing variable certainty in
-frequency data}
+\section*{Representing rare variants in frequency data}
 
-One under-appreciated problem with allele frequency maps is that not all
-data points have equal levels of certainty. For some locations, sample
-sizes are small, and the reported allele frequency may be quite far from
-the true population frequency due to sampling error. To address this
-issue, we use varying transparency in a population's pie chart: sample
-allele frequencies with higher levels of sampling error will be made
-more transparent, and hence less visible on the map {[}Figure 3{]}.
+An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset.
+This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays, which biased towards variants that are common in human populations (5-50 \% in minor allele frequency).
+With the advent of next generation sequencing technologies, novel array designs focusing on rarer variants, and large samples of thousands of individuals, it is now common for datasets to contain rare variants \cite{10002015global, nelson2012abundance, tennessen2012evolution}.
+In current visualization schemes, such as the HGDP Selection Browser, rare variants would be represented as narrow slivers in a pie chart, nearly invisible to the naked eye.
 
-\begin{CodeChunk}
-\begin{figure}
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=.75\textwidth]{figures/pdf/figure4.pdf}
+ \caption{Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
+The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.
+The data are from \cite{10002015global}.}
+\end{figure}
 
-{\centering \includegraphics{paper_files/figure-latex/figure3-1}
+To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
+Specifically, we  use a “frequency scale” that is indicated in a legend below the map, and redundantly by using color, that indicates the scale for the frequencies displayed [Figure 4].
+Much like scientific notation, this allows a wide range of frequencies to be displayed.
+[xxJN: I'd like us to use the example maps, plus a figure that shows a table of the colors being used and how they translate.
+I have a figure I've used in talks that should work.
+xxJM: see latex table in progress]
 
-}
+\begin{figure}[h]
+ \centering
+ \includegraphics[width=\textwidth]{figures/pdf/figure5.pdf}
+ \caption{Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
+The left map shows the original population locations while the right shows the application of the force directed layout.}
+\end{figure}
 
-\caption[Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies]{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
-The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.}\label{fig:figure3}
-\end{figure}
-\end{CodeChunk}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section*{Representing rare variants in frequency
-data}\label{representing-rare-variants-in-frequency-data}
-\addcontentsline{toc}{section}{Representing rare variants in frequency
-data}
+\section*{Additional features of the interface}
 
-An additional challenge is that allele frequencies between variants
-often differ greatly, sometimes by orders of magnitude in a single
-dataset. This has not been an pervasive problem until recently, as most
-population genetic samples were based on genotype arrays, biased towards
-variants that are common in human populations (5-50\% in minor allele
-frequency). With the advent of next generation sequencing technologies
-and large samples of thousands of individuals, it is common for datasets
-to contain rare variants {[}17{]}, {[}4{]}, {[}22{]}.
+In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
+To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure 5].
+Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
 
-\begin{CodeChunk}
-\begin{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-{\centering \includegraphics{paper_files/figure-latex/figure4-1}
+\section*{Access to the underlying frequency data}
 
-}
+To provide an interface to the population minor allele frequency data stored in the SQL database we use a REST API implemented in the python library Flask-RESTful @citeXXX.
+The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
+For the front-end, GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g.
+latitude, longitude, population label, sample size, and frequency scale.
+Genetic variants can be queried by chromosome position, rsid, or randomly.
+Example HTTP requests and json response can be seen in Examples 1,2,3 (below).
 
-\caption[Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps]{Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
-The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.}\label{fig:figure4}
-\end{figure}
-\end{CodeChunk}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-In current visualization schemes, such as the HGDP Selection Browser,
-rare variants would be represented as narrow slivers in a pie chart,
-nearly invisible to the naked eye. To address this challenge we re-scale
-frequencies for rare variants, so that small frequencies become visible.
-Specifically, we will use a ``frequency scale'' that is indicated below
-the map, and redundantly using color, that will indicate a constant
-scale for the frequencies displayed {[}Figure 4{]}. Much like scientific
-notation, this allows a wide range of frequencies to be displayed{]}.
+\section*{Caveats}
 
-\section*{Interface features}\label{interface-features}
-\addcontentsline{toc}{section}{Interface features}
+A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.
+While doing so is generally immensely helpful, it has inherent complexity and limitations.
+For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g.
+the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
+as determined in practice by the birthplaces of recent ancestors, such as grandparents).
+We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that is based on those used by the initial analysts of the data.
+A future feature will allow alternative location schema to be used for the populations in a dataset.
 
-In many datasets where populations are sampled densely in geographic
-space, one problem is that allele frequency plots begin to overlap each
-other and obscure information. To address this issue, we use
-force-directed layouts of the populations such that no two points are
-overlapping each other, and yet the points will be pulled towards their
-true origins. Lines anchoring the points visibly to their original
-sampling locations will be used to make sure true sampling locations are
-indicated {[}Figure 5{]}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{CodeChunk}
-\begin{figure}
+\section*{Conclusion}
 
-{\centering \includegraphics{paper_files/figure-latex/figure5-1}
+By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
+We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
 
-}
+We also envision a variety of future extensions to the GGV that would allow for further dissection of geographical structure in large-scale population genomic datasets.
+Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP which would be widely relevant to medical geneticist conducting genome-wide association studies as well as population geneticists interested in signatures of selection.
+We imagine incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators a like.
 
-\caption[Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position]{Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
-The left map shows the original population locations while the right shows the application of the force directed layout.}\label{fig:figure5}
-\end{figure}
-\end{CodeChunk}
+The constellation of common single-nucleotide polymorphism (SNP) alleles carried by an individual can provide some insight to the geographic origins of their recent ancestors.
+More recently, the increasing availability of data on rare variants is opening up new opportunities for ancestry inference.
+Rare variants are particularly informative for ancestry as they tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events \cite{nelson2012abundance, mathieson2014demography}.
+By assessing the geographic distribution of rare variants carried by an individual one can gain information about their ancestry (e.g.
+\cite{BrownPasniuc2014PloSCompBio, KennewickManReport}).
+In these applications, geographic maps of the genetic variants an individual carries can be useful and informative.
+We imagine developing tools to upload personal genetic data from companies such as from 23andMe or AncestryDNA would allow users to visualize the set of rare variants they carry and could be of wide interest to the public.
+In order to achieve this goal we plan to extend the ability of the GGV API to query and return data from thousands of genetic variants at a time.
 
-\section*{Underlying Frequency API}\label{underlying-frequency-api}
-\addcontentsline{toc}{section}{Underlying Frequency API}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-To interface population minor allele frequency data stored in the SQL
-database with the front-end D3.js visualization we use a REST API
-implemented in the python library Flask-RESTful \textbf{???}. GET
-requests are submitted to the database via HTTP, returning json
-formatted allele frequency data and the meta-data associated with each
-population and genetic variant e.g.~latitude, longitude, population
-label, sample size, and frequency scale. Genetic variants can be queried
-by chromosome position, rsid, or randomly. Example HTTP requests and
-json response can be seen below:
+\section*{Acknowledgements}
 
-\subsection*{Query by rsid}\label{query-by-rsid}
-\addcontentsline{toc}{subsection}{Query by rsid}
+Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank members of the Novembre Lab for supportive conversations.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection*{Example 1: Query by rsid}
 
@@
 
-\subsection*{Query by chromosome
-position}\label{query-by-chromosome-position}
-\addcontentsline{toc}{subsection}{Query by chromosome position}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection*{Example 2: Query by chromosome position}
 
@@
 
-\subsection*{Random query}\label{random-query}
-\addcontentsline{toc}{subsection}{Random query}
+\subsection*{Example 3: Random query}
 
@@
 
-\section*{Caveats}\label{caveats}
-\addcontentsline{toc}{section}{Caveats}
+\subsection*{xxJM: Freq scale table in Progress}
 
-A major challenge of using a geographic representation of genetic
-variation in humans is that the samples must be associated with a
-geographic location. While doing so is generally immensely helpful, it
-has inherent complexity. For example, practitioners must make choices
-regarding representing where an individual was sampled for the study
-(e.g.~the city of a major research center) or choosing a location that
-is more representative of an individual's ancestral origins (e.g.~as
-determined in practice by the birthplaces of recent ancestors, such as
-grandparents). We do not proscribe a general solution to this problem
-and instead use a set of locations for each dataset that aligns with
-those used by the initial analysts of the data. A future feature will be
-to allow alternative location schema to be used for the populations in a
-dataset.
+\begin{center}
+  \begin{table}[h]
+    \begin{tabular}
+        {lllll} \hline \textbf{Frequency Scale} & \textbf{Frequency} & \textbf{Pie} \\
+        \hline 1 & .25 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_blue.pdf}} \\
+        .1 & .025 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_red.pdf}} \\
+        .01 & .0025 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}} 
+    \end{tabular}
+  \end{table}
+\end{center}
 
-\section*{Conclusion}\label{conclusion}
-\addcontentsline{toc}{section}{Conclusion}
-
-By allowing rapid generation of allele frequency maps, we hope to
-facilitate the interpretation of variant function and history by
-practicing geneticists. We also hope the ability to query random
-variants from major human population genetic samples will allow students
-to appreciate the structure of human genetic diversity in a more
-approachable and intuitive form than alternative visualizations.
-
-\section*{Acknowledgements}\label{acknowledgements}
-\addcontentsline{toc}{section}{Acknowledgements}
-
-Support for this work was provided by the NIH-BD2K initiative (1U01
-CA198933-0). The authors would also like to thank members of the
-Novembre Lab for supportive conversations.
-
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\setlength\parindent{0pt}
-
-\hypertarget{refs}{}
-\hypertarget{ref-rosenberg2010genome}{}
-1. Rosenberg NA, Huang L, Jewett EM, Szpiech ZA, Jankovic I, Boehnke M.
-Genome-wide association studies in diverse populations. Nature Reviews
-Genetics. Nature Publishing Group; 2010;11: 356--366.
-
-\hypertarget{ref-novembre2009spatial}{}
-2. Novembre J, Di Rienzo A. Spatial patterns of variation due to natural
-selection in humans. Nature Reviews Genetics. Nature Publishing Group;
-2009;10: 745--755.
-
-\hypertarget{ref-coop2010using}{}
-3. Coop G, Witonsky D, Di Rienzo A, Pritchard JK. Using environmental
-correlations to identify loci underlying local adaptation. Genetics.
-Genetics Soc America; 2010;185: 1411--1423.
-
-\hypertarget{ref-nelson2012abundance}{}
-4. Nelson MR, Wegmann D, Ehm MG, Kessner D, Jean PS, Verzilli C, et al.
-An abundance of rare functional variants in 202 drug target genes
-sequenced in 14,002 people. Science. American Association for the
-Advancement of Science; 2012;337: 100--104.
-
-\hypertarget{ref-mathieson2014demography}{}
-5. Mathieson I, McVean G. Demography and the age of rare variants. PLoS
-Genet. Public Library of Science; 2014;10: e1004528.
-
-\hypertarget{ref-price2006principal}{}
-6. Price AL, Patterson NJ, Plenge RM, Weinblatt ME, Shadick NA, Reich D.
-Principal components analysis corrects for stratification in genome-wide
-association studies. Nature genetics. Nature Publishing Group; 2006;38:
-904--909.
-
-\hypertarget{ref-patterson2006population}{}
-7. Patterson N, Price AL, Reich D. Population structure and
-eigenanalysis. PLoS genet. Public Library of Science; 2006;2: e190.
-
-\hypertarget{ref-pritchard2000inference}{}
-8. Pritchard JK, Stephens M, Donnelly P. Inference of population
-structure using multilocus genotype data. Genetics. Genetics Soc
-America; 2000;155: 945--959.
-
-\hypertarget{ref-alexander2009fast}{}
-9. Alexander DH, Novembre J, Lange K. Fast model-based estimation of
-ancestry in unrelated individuals. Genome research. Cold Spring Harbor
-Lab; 2009;19: 1655--1664.
-
-\hypertarget{ref-furchedata}{}
-10. Furche T, Gottlob G, Libkin L, Orsi G, Paton NW. Data wrangling for
-big data: Challenges and opportunities.
-
-\hypertarget{ref-rajeevan2011alfred}{}
-11. Rajeevan H, Soundararajan U, Kidd JR, Pakstis AJ, Kidd KK. ALFRED:
-An allele frequency resource for research and teaching. Nucleic acids
-research. Oxford Univ Press; 2011; gkr924.
-
-\hypertarget{ref-pickrell2009signals}{}
-12. Pickrell JK, Coop G, Novembre J, Kudaravalli S, Li JZ, Absher D, et
-al. Signals of recent positive selection in a worldwide sample of human
-populations. Genome research. Cold Spring Harbor Lab; 2009;19: 826--837.
-
-\hypertarget{ref-wessel2013generic}{}
-13. Wessel P, Smith WH, Scharroo R, Luis J, Wobbe F. Generic mapping
-tools: Improved version released. Eos, Transactions American Geophysical
-Union. Wiley Online Library; 2013;94: 409--410.
-
-\hypertarget{ref-coop2009role}{}
-14. Coop G, Pickrell JK, Novembre J, Kudaravalli S, Li J, Absher D, et
-al. The role of geography in human adaptation. PLoS Genet. Public
-Library of Science; 2009;5: e1000500.
-
-\hypertarget{ref-dudley2013exploring}{}
-15. Dudley JT, Karczewski KJ. Exploring personal genomics. Oxford
-University Press; 2013.
-
-\hypertarget{ref-li2008worldwide}{}
-16. Li JZ, Absher DM, Tang H, Southwick AM, Casto AM, Ramachandran S, et
-al. Worldwide human relationships inferred from genome-wide patterns of
-variation. science. American Association for the Advancement of Science;
-2008;319: 1100--1104.
-
-\hypertarget{ref-10002015global}{}
-17. Consortium 1GP, others. A global reference for human genetic
-variation. Nature. Nature Publishing Group; 2015;526: 68--74.
-
-\hypertarget{ref-meyer2012high}{}
-18. Meyer M, Kircher M, Gansauge M-T, Li H, Racimo F, Mallick S, et al.
-A high-coverage genome sequence from an archaic denisovan individual.
-Science. American Association for the Advancement of Science; 2012;338:
-222--226.
-
-\hypertarget{ref-lazaridis2014ancient}{}
-19. Lazaridis I, Patterson N, Mittnik A, Renaud G, Mallick S, Kirsanow
-K, et al. Ancient human genomes suggest three ancestral populations for
-present-day europeans. Nature. Nature Publishing Group; 2014;513:
-409--413.
-
-\hypertarget{ref-bostock2011d3}{}
-20. Bostock M, Ogievetsky V, Heer J. D\(^3\) data-driven documents.
-Visualization and Computer Graphics, IEEE Transactions on. IEEE;
-2011;17: 2301--2309.
-
-\hypertarget{ref-sherry2001dbsnp}{}
-21. Sherry ST, Ward M-H, Kholodov M, Baker J, Phan L, Smigielski EM, et
-al. DbSNP: The nCBI database of genetic variation. Nucleic acids
-research. Oxford Univ Press; 2001;29: 308--311.
-
-\hypertarget{ref-tennessen2012evolution}{}
-22. Tennessen JA, Bigham AW, O'Connor TD, Fu W, Kenny EE, Gravel S, et
-al. Evolution and functional impact of rare coding variation from deep
-sequencing of human exomes. Science. American Association for the
-Advancement of Science; 2012;337: 64--69.
+\clearpage
+\bibliographystyle{genetics}
+\bibliography{paper}
 

### 2016-07-16 paper.tex
changed some grammer in conclusions

@@
 We also envision a variety of future extensions to the GGV that would allow for further dissection of geographical structure in large-scale population genomic datasets.
-Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP which would be widely relevant to medical geneticist conducting genome-wide association studies as well as population geneticists interested in signatures of selection.
-We imagine incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators a like.
+Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP.
+This feature would be relevant to both medical geneticist conducting genome-wide association studies with in inteterests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
+We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators alike.
 

### 2016-07-21 paper.tex
Change citation style; minor style/typo edits; minor restructuring

@@
 what population should be studied to observe variant carriers?); or 3) the biology of the variant in question (e.g.
-does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?)  \cite{rosenberg2010genome, novembre2009spatial, coop2010using}.
+does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?)  \citep{rosenberg2010genome, novembre2009spatial, coop2010using}.
 A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
@@
 Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets.
-For example, is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
-One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \cite{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \cite{pritchard2000inference} or ADMIXTURE \cite{alexander2009fast}.
+For example, it is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
+One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \citep{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \citep{pritchard2000inference} or ADMIXTURE \citep{alexander2009fast}.
 While these approaches are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
@@
 
-Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods \cite{furchedata} and map making techniques that are unfamiliar to most.
+Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods \citep{furchedata} and map making techniques that are unfamiliar to most.
 Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.
 
-Our work is inspired by past tools such as the ALFRED database \cite{rajeevan2011alfred} and the maps available on the HGDP Selection browser \cite{pickrell2009signals}.
-One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \cite{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
-The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g \cite{pickrell2009signals}, \cite{coop2009role}, books e.g.
-\cite{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
-Reference datasets for population genetic variation have greatly expanded since the release of the HGDP Illumina 650Y dataset \cite{li2008worldwide} that formed the basis of the HGDP Selection Browser maps.
-The most notable advance is the publication of the 1000 Genomes Phase 3 data \cite{10002015global} though additional datasets are continually coming online \cite{meyer2012high, lazaridis2014ancient}.
+Our work is inspired by past tools such as the ALFRED database \citep{rajeevan2011alfred} and the maps available on the HGDP Selection browser \citep{pickrell2009signals}.
+One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \citep{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
+The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles \citep[e.g.]{pickrell2009signals,coop2009role}, books \citep[e.g.]{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
+
+Reference datasets for population genetic variation have greatly expanded since the release of the HGDP Illumina 650Y dataset \citep{li2008worldwide} that formed the basis of the HGDP Selection Browser maps.
+The most notable advance is the publication of the 1000 Genomes Phase 3 data \citep{10002015global} though additional datasets are continually coming online \citep{meyer2012high, lazaridis2014ancient}.
 In addition, novel approaches for data visualization have become more widely available.
-In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer useful methods for interactivity, the advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing \cite{bostock2011d3}.
+In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer useful methods for interactivity, the advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing \citep{bostock2011d3}.
 
@@
 These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.}
+The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented using the python library flask.}
 \end{figure}
@@
 
-The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \cite{bostock2011d3} to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
+The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \citep{bostock2011d3} to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
+The front-end provides legends for the map and various configuration boxes to allow users to query different datasets or choose visualization options.
 
-In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \cite{10002015global}, Human Genome Diversity project \cite{li2008worldwide}, we have developed an SQL database and REST API for querying allele frequencies by chromosome and position, by reference SNP identifier \cite{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1].
-
-The front-end of the project is a browser-based visualization application that displays the geographic map of a particular variant and provides legends for the map and various boxes to allow users to query different datasets or choose visualization options.
+In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global}, Human Genome Diversity project \citep{li2008worldwide}, we have developed an SQL database and REST API \citep{SQLrestAPI} for querying allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1].
 While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
@@
  \centering
- \includegraphics[width=.55\textwidth]{figures/png/figure2.png}
- \caption{Example screenshot from the Geography of Genetic Variants browser using the \cite{10002015global} data.
+ \includegraphics[width=.95\textwidth]{figures/png/figure2.png}
+ \caption{Example screenshot from the Geography of Genetic Variants browser using the \citep{10002015global} data.
 Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}
@@
 After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies.
-Pie charts are displayed as points positioned at the population's region of origin and the projection / map-view is chosen based off of the geographic configuration of populations in a given dataset [Figure 2].
+Pie charts are displayed as points positioned at the population's region of origin and the geographic range displayed is chosen based off of the geographic configuration of populations in a given dataset [Figure 2].
 
@@
 To address this issue, we use varying transparency in a population's pie chart: sampled populations with higher levels of sampling error (i.e.
-$n < 30$) are made more transparent, and hence less visible on the map [Figure 3].
+$n < 30$) are made more transparent, and hence less visible, on the map [Figure 3].
 
@@
  \centering
- \includegraphics[width=.65\textwidth]{figures/pdf/figure3.pdf}
+ \includegraphics[width=.5\textwidth]{figures/pdf/figure3.pdf}
  \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
 The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.
-The frequency data and population identifiers are from \cite{novembre2008genes}.}
+The frequency data and population identifiers are from \citep{novembre2008genes}.}
 \end{figure}
@@
 An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset.
-This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays, which biased towards variants that are common in human populations (5-50 \% in minor allele frequency).
-With the advent of next generation sequencing technologies, novel array designs focusing on rarer variants, and large samples of thousands of individuals, it is now common for datasets to contain rare variants \cite{10002015global, nelson2012abundance, tennessen2012evolution}.
-In current visualization schemes, such as the HGDP Selection Browser, rare variants would be represented as narrow slivers in a pie chart, nearly invisible to the naked eye.
+This has not been a pervasive problem until recently, as most population genetic samples were genotyped on SNP arrays, which have been biased towards variants that are common in human populations (5-50 \% in minor allele frequency).
+With the combination of next generation sequencing technologies, new array designs focusing on rarer variants, and large samples of thousands of individuals, it is now routine for datasets to contain rare variants \citep{10002015global, nelson2012abundance, tennessen2012evolution}.
+A new challenge is that in visualization schemes using proportional area to represent frequency (such as standard pie charts), rare variants would be represented as narrow slivers, nearly invisible to the naked eye.
 
@@
 The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.
-The data are from \cite{10002015global}.}
+The data are from \citep{10002015global}.}
 \end{figure}
@@
 Specifically, we  use a “frequency scale” that is indicated in a legend below the map, and redundantly by using color, that indicates the scale for the frequencies displayed [Figure 4].
-Much like scientific notation, this allows a wide range of frequencies to be displayed.
-[xxJN: I'd like us to use the example maps, plus a figure that shows a table of the colors being used and how they translate.
-I have a figure I've used in talks that should work.
-xxJM: see latex table in progress]
+Much like scientific notation, this allows a wide range of frequencies to be displayed (Table \ref{tab:rare}).
+
+\begin{center}
+  \begin{table}[h]
+    \begin{tabular}
+        {lllll} \hline  \textbf{Sample Frequency}& \textbf{Display Frequency Scale} & \textbf{Displayed Image} \\
+        \hline 0.25 & 1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_blue.pdf}} \\
+        0.025 &  0.1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_red.pdf}} \\
+        0.0025 &  0.01 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}} \\
+        0.00025 &  0.001 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}}
+    \end{tabular}
+  \caption{Rare variants present a challenge for display.
+To address this challenge, the GGV browser changes the displayed image and the frequency scale of the map depending on the input sample frequency.
+As an example, a variant with a frequency 0.0025 is shown as a pie-chart that is 25\% full and a frequency scale of 0.01 is marked in the legend of the map.}
+  \label{tab:rare}
+  \end{table}
+
+\end{center}
 
@@
 
-\section*{Caveats}
+\section*{Discussion}
+
+By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
+We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
 
@@
 as determined in practice by the birthplaces of recent ancestors, such as grandparents).
-We do not proscribe a general solution to this problem and instead use a set of locations for each dataset that is based on those used by the initial analysts of the data.
+We do not proscribe a general solution to this problem, and for the current defaults we use locations based on the approach taken in the source publications.
 A future feature will allow alternative location schema to be used for the populations in a dataset.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Conclusion}
-
-By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
-We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
 
@@
 Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP.
-This feature would be relevant to both medical geneticist conducting genome-wide association studies with in inteterests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
+This feature would be relevant to both medical geneticists conducting genome-wide association studies with in inteterests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
 We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators alike.
 
-The constellation of common single-nucleotide polymorphism (SNP) alleles carried by an individual can provide some insight to the geographic origins of their recent ancestors.
+%The constellation of common single-nucleotide polymorphism (SNP) alleles carried by an individual can provide some insight to the geographic origins of their recent ancestors.
 More recently, the increasing availability of data on rare variants is opening up new opportunities for ancestry inference.
-Rare variants are particularly informative for ancestry as they tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events \cite{nelson2012abundance, mathieson2014demography}.
+Rare variants are particularly informative for ancestry as they tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events \citep{nelson2012abundance, mathieson2014demography}.
 By assessing the geographic distribution of rare variants carried by an individual one can gain information about their ancestry (e.g.
-\cite{BrownPasniuc2014PloSCompBio, KennewickManReport}).
+\citep{BrownPasniuc2014PloSCompBio, KennewickManReport}).
 In these applications, geographic maps of the genetic variants an individual carries can be useful and informative.
@@
 
-\subsection*{xxJM: Freq scale table in Progress}
-
-\begin{center}
-  \begin{table}[h]
-    \begin{tabular}
-        {lllll} \hline \textbf{Frequency Scale} & \textbf{Frequency} & \textbf{Pie} \\
-        \hline 1 & .25 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_blue.pdf}} \\
-        .1 & .025 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_red.pdf}} \\
-        .01 & .0025 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}} 
-    \end{tabular}
-  \end{table}
-\end{center}
-
 \clearpage

### 2016-07-28 paper.tex
fixed figure3, added purple pie, and edited some of the citations

@@
 Collectively the geographic distribution of many genetic variants can inform on how populations have been related through time (e.g.~levels of gene flow and divergence).
-The distribution of a genetic variant can also be informative background knowledge for medical/clinical geneticists.
+The distribution of a genetic variant can also be informative background knowledge for medical / clinical geneticists.
 As a result, visual inspection of geographic maps for genetic variants is common practice in genetic studies.
@@
 One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \citep{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
-The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles \citep[e.g.]{pickrell2009signals,coop2009role}, books \citep[e.g.]{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).
+The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles \citep[e.g.]{pickrell2009signals,coop2009role}, books \citep[e.g.]{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser) \citep{kent2002human}.
 
@@
  \centering
- \includegraphics[width=.5\textwidth]{figures/pdf/figure3.pdf}
+ \includegraphics[width=.5\textwidth]{figures/png/figure3.png}
  \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
@@
         \hline 0.25 & 1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_blue.pdf}} \\
-        0.025 &  0.1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_red.pdf}} \\
-        0.0025 &  0.01 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}} \\
-        0.00025 &  0.001 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}}
+        0.025 &  0.1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}} \\
+        0.0025 &  0.01 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_red.pdf}} \\
+        0.00025 &  0.001 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_purple.pdf}}
     \end{tabular}
@@
 This feature would be relevant to both medical geneticists conducting genome-wide association studies with in inteterests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
-We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators alike.
+We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators alike \citep{kent2002human}.
 

### 2016-08-04 cabios-template/main.tex
almost done with citations!

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[short Title]{Visualizing the Geography of Genetic Variants}
\author[Sample \textit{et~al}.]{Corresponding Author\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
$^{\text{\sf 2}}$Department, Institution, City, Post Code,
Country.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text
Text Text Text Text Text.\\
\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text. Figure~\ref{fig:01}
shows that the above method  Text Text Text Text  Text Text Text
Text Text Text  Text Text.  \citep{Bag01} wants to know about
{\ldots}{\ldots} text follows.
\begin{equation}
\sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
\end{equation}

%\enlargethispage{12pt}

\section{Approach}

Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
.....

\begin{methods}
\section{Methods}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text
text\vspace*{1pt}

\begin{itemize}
\item for bulleted list, use itemize
\item for bulleted list, use itemize
\item for bulleted list, use itemize\vspace*{1pt}
\end{itemize}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
Text Text Text Text Text Text Text Text Text Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text\vadjust{\newpage}.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
Text Text Text Text Text Text Text Text Text Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\subsection{This is subheading}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text
text. Text Text Text Text Text Text  Text Text Text Text Text Text
Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\subsubsection{This is subsubheading}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text\break
text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text
textText Text Text Text Text Text  Text Text Text Text Text Text
Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

\enlargethispage{6pt}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\begin{table}[!t]
\processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule head1 &
head2 & head3 & head4\\\midrule
row1 & row1 & row1 & row1\\
row2 & row2 & row2 & row2\\
row3 & row3 & row3 & row3\\
row4 & row4 & row4 & row4\\\botrule
\end{tabular}}{This is a footnote}
\end{table}

\end{methods}

\begin{figure}[!tpb]%figure1
\fboxsep=0pt\colorbox{gray}{\begin{minipage}[t]{235pt} \vbox to 100pt{\vfill\hbox to
235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}\vfill}
\end{minipage}}
%\centerline{\includegraphics{fig01.eps}}
\caption{Caption, caption.}\label{fig:01}
\end{figure}

%\begin{figure}[!tpb]%figure2
%%\centerline{\includegraphics{fig02.eps}}
%\caption{Caption, caption.}\label{fig:02}
%\end{figure}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\subsection{Test1}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\section{Discussion}

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text.

Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text
Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
the above method Text Text. Text Text Text  Text Text Text Text
Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
method Text Text. Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
Text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text
Text Text Text Text Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text
Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about  text text text text

\begin{enumerate}
\item this is item, use enumerate
\item this is item, use enumerate
\item this is item, use enumerate
\end{enumerate}

Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
above method  Text Text Text Text Text Text Text Text Text Text
Text Text.  \citealp{Boffelli03} might want to know about text
text text text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text
Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text Text Text Text Text Text\break Text.

Text Text Text Text Text Text  Text Text Text Text Text Text Text
Text Text  Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text\vspace*{-10pt}

\section*{Acknowledgements}

Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text
text text text\vspace*{-12pt}

\section*{Funding}

This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\begin{thebibliography}{}

\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Bag {\it et~al}., 2001]{Bag01}
Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
in nigral dopaminergic neurnol cell: correlation with the known
pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
Res.}, \textbf{110}(Suppl. 1), 76--84.

\bibitem[Lehmann, 1986]{Leh86}
Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

\bibitem[Crenshaw and Jones, 2003]{Cre03}
Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
cancer management: one tumor, one chip. \textit{Bioinformatics},
doi:10.1093/bioinformatics/btn000.

\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
???--???.

\bibitem[Bardet, 1920]{Bar20}
Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
polydactylie et retinite pigmentaire (contribution a l'etude des
formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
institution, Paris, France.

\end{thebibliography}
\end{document}

### 2016-08-04 paper.tex
almost done with citations!

@@
 
-In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global}, Human Genome Diversity project \citep{li2008worldwide}, we have developed an SQL database and REST API \citep{SQLrestAPI} for querying allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1].
+In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global}, Human Genome Diversity project \citep{li2008worldwide}, we have developed an SQL database and REST API \citep{grinberg2014flask} for querying allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1].
 While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
@@
 
-To provide an interface to the population minor allele frequency data stored in the SQL database we use a REST API implemented in the python library Flask-RESTful @citeXXX.
+To provide an interface to the population minor allele frequency data stored in the SQL database we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}.
 The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.

### 2016-08-04 paper.tex
added flat icon citations

@@
 \usepackage[labelfont=bf]{caption}
+\usepackage[hang,flushmargin]{footmisc} 
 \makeatletter
@@
 These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented using the python library flask.}
+The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented using the python library flask.
+}
 \end{figure}
@@
 
-In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global}, Human Genome Diversity project \citep{li2008worldwide}, we have developed an SQL database and REST API \citep{grinberg2014flask} for querying allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1].
+In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global}, Human Genome Diversity project \citep{li2008worldwide}, we have developed an SQL database and REST API \citep{grinberg2014flask} for querying allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1]\footnote{Database icon made by http://www.flaticon.com/authors/madebyoliver from www.flaticon.com and monitor icon made by http://www.flaticon.com/authors/gregor-cresnar from www.flaticon.com}.
 While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.

### 2016-08-05 paper.tex
Cleaning up bibliography

@@
 Our work is inspired by past tools such as the ALFRED database \citep{rajeevan2011alfred} and the maps available on the HGDP Selection browser \citep{pickrell2009signals}.
-One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \citep{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
-The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles \citep[e.g.]{pickrell2009signals,coop2009role}, books \citep[e.g.]{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser) \citep{kent2002human}.
+One of us (JN) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \citep{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
+The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser) \citep{kent2002human}.
 
@@
 Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP.
-This feature would be relevant to both medical geneticists conducting genome-wide association studies with in inteterests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
-We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser within the GGV would be greatly utilized by researchers and educators alike \citep{kent2002human}.
+This feature would be relevant to both medical geneticists conducting genome-wide association studies with in interests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
+We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser \citep{kent2002human} within the GGV would be greatly utilized by researchers and educators alike.
 

### 2016-08-08 paper.tex
Polishing...

@@
 \author[1]{Joseph H. Marcus$^{*}$}
-\author[1]{John Novembre\thanks{Address correspondence to JM (\texttt{josephhmarcus@gmail.com}) or JN (\texttt{jnovembre@uchicago.edu}).}}
+\author[1]{John Novembre\thanks{Address correspondence to JHM (\texttt{jhmarcus@uchicago.edu}) or JN (\texttt{jnovembre@uchicago.edu}).}}
 \affil[1]{\small{Department of Human Genetics, University of Chicago, Chicago, IL,USA}}
@@
 \begin{abstract}
-One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its global frequency, is its geographic distribution.
-The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it has spread to, and in turn how migration, genetic drift, and natural selection have acted on the variant.
-Collectively the geographic distribution of many genetic variants can inform on how populations have been related through time (e.g.~levels of gene flow and divergence).
-The distribution of a genetic variant can also be informative background knowledge for medical / clinical geneticists.
+One of the key characteristics of any genetic variant is its geographic distribution.
+The geographic distribution can shed light on where an allele first arose, what populations it has spread to, and in turn on how migration, genetic drift, and natural selection have acted.
+The distribution of a genetic variant can also be of great utility for medical/clinical geneticists.
+Collectively the geographic distribution of many genetic variants can reveal population structure.
 As a result, visual inspection of geographic maps for genetic variants is common practice in genetic studies.
-Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants.
-Through an efficient SQL database, REST API and dynamic front-end the \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}} browser rapidly provides maps of allele frequencies in populations distributed across the globe.
+Here we develop an interactive visualization tool for rapidly displaying the geographic distribution of genetic variants.
+Through a REST API and dynamic front-end the \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}} browser provides maps of allele frequencies in populations distributed across the globe.
 \end{abstract}
-\newpage
 
@@
 
-Genetics researchers often face the problem that they have identified one or many genetic variants of interest (e.g.
-from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant.
-For example, the researcher may hope to address: 1) implications for genomic medicine (e.g.
-is a risk allele geographically localized to a certain patient population?); 2) design of follow-up studies (e.g.
-what population should be studied to observe variant carriers?); or 3) the biology of the variant in question (e.g.
-does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?)  \citep{rosenberg2010genome, novembre2009spatial, coop2010using}.
+Genetics researchers often face the problem that they have identified one or many genetic variants of interest using an approach such as a genome-wide association study and then would like to know the geographic distribution of the variant.
+For example, the researcher may hope to address: 1) implications for genomic medicine \citep[e.g.
+Is a risk allele geographically localized to a certain patient population?
+What population should be studied to observe variant carriers?][]{rosenberg2010genome}; or 2) the evolutionary history of the variant in question \citep[e.g.
+does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?][]{novembre2009spatial,coop2010using}.
 A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
 
-Contemporary population genetics researchers are also faced with the challenge of large, high dimensional datasets.
+Contemporary population genetics researchers are also faced with the challenge of large, high-dimensional datasets.
 For example, it is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
@@
 
-Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data wrangling methods \citep{furchedata} and map making techniques that are unfamiliar to most.
+Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data-wrangling methods \citep{kandel2011research} and map-making techniques that are unfamiliar to most.
 Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.
@@
 One of us (JN) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \citep{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
-The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been ported and made available on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser) \citep{kent2002human}.
+The plots from the HGDP Selection Browser have proved useful, have appeared in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been made available on the UCSC Genome Browser \citep[available under the HGDP Allele Freq track of the browser][]{kent2002human}.
 
 Reference datasets for population genetic variation have greatly expanded since the release of the HGDP Illumina 650Y dataset \citep{li2008worldwide} that formed the basis of the HGDP Selection Browser maps.
-The most notable advance is the publication of the 1000 Genomes Phase 3 data \citep{10002015global} though additional datasets are continually coming online \citep{meyer2012high, lazaridis2014ancient}.
+The most notable advance is the publication of the 1000 Genomes Phase 3 data \citep{10002015global} though additional datasets are continually coming online \citep[e.g.][]{lazaridis2014ancient}.
 In addition, novel approaches for data visualization have become more widely available.
@@
 
-\begin{figure}[h]
- \centering
- \includegraphics[width=.4\textwidth]{figures/pdf/figure1.pdf}
- \caption{Schematic diagram of the Geography of Genetic Variants (GGV) browser.
+%\begin{figure}[h]
+% \centering
+% \includegraphics[width=.4\textwidth]{figures/pdf/figure1.pdf}
+% \caption{Schematic diagram of the Geography of Genetic Variants (GGV) browser.
 We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
@@
 }
-\end{figure}
+%\end{figure}
 
@@
 
-In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global}, Human Genome Diversity project \citep{li2008worldwide}, we have developed an SQL database and REST API \citep{grinberg2014flask} for querying allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs [Figure 1]\footnote{Database icon made by http://www.flaticon.com/authors/madebyoliver from www.flaticon.com and monitor icon made by http://www.flaticon.com/authors/gregor-cresnar from www.flaticon.com}.
-While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
-For instance, in data with deep population subdivision, it is obvious in the consistent patterns of differentiation observed across markers.
-We also expect this will be useful in teaching contexts – as it provides a highly visual way for learners to understand human genetic variation.
+In order to allow for easy access to commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global} or Human Genome Diversity project \citep{li2008worldwide}, we have developed a REST API \citep{grinberg2014flask} for accessing data.
+The API allows querying of allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs.
+While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
+We find this to be especially useful in teaching contexts, as it provides a highly visual way for learners to understand human genetic variation.
 
@@
  \includegraphics[width=.95\textwidth]{figures/png/figure2.png}
- \caption{Example screenshot from the Geography of Genetic Variants browser using the \citep{10002015global} data.
+ \caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
 Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}
+\label{fig:screenshot}
 \end{figure}
 
-After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies.
-Pie charts are displayed as points positioned at the population's region of origin and the geographic range displayed is chosen based off of the geographic configuration of populations in a given dataset [Figure 2].
+After a query, the GGV displays the allele frequencies for a set of populations as a collection of pie charts where each represents the minor and major allele frequency in a single population.
+Pie charts are displayed as points at a latitude and longitude assoicated with a population and the map boundaries are chosen based off of the geographic configuration of populations in a given dataset [Figure \ref{fig:screenshot}].
 
@@
 For some locations, sample sizes are small, and the reported allele frequency may be quite far from the true population frequency due to sampling error.
-To address this issue, we use varying transparency in a population's pie chart: sampled populations with higher levels of sampling error (i.e.
-$n < 30$) are made more transparent, and hence less visible, on the map [Figure 3].
+To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:uncertainty}].
 
@@
  \centering
- \includegraphics[width=.5\textwidth]{figures/png/figure3.png}
+ \includegraphics[width=.8\textwidth]{figures/png/figure3.png}
  \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
 The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.
-The frequency data and population identifiers are from \citep{novembre2008genes}.}
+The frequency data and population identifiers are from \citet{novembre2008genes}.}
+\label{fig:uncertainty}
 \end{figure}
@@
 This has not been a pervasive problem until recently, as most population genetic samples were genotyped on SNP arrays, which have been biased towards variants that are common in human populations (5-50 \% in minor allele frequency).
-With the combination of next generation sequencing technologies, new array designs focusing on rarer variants, and large samples of thousands of individuals, it is now routine for datasets to contain rare variants \citep{10002015global, nelson2012abundance, tennessen2012evolution}.
-A new challenge is that in visualization schemes using proportional area to represent frequency (such as standard pie charts), rare variants would be represented as narrow slivers, nearly invisible to the naked eye.
+With the combination of next generation sequencing technologies, new array designs focusing on rarer variants, and studies with thousands of individuals or more, it is now routine for the majority of variants to be rare \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}.
+In visualization schemes using proportional area to represent frequency (such as standard pie charts), rare variants would be represented as narrow slivers, nearly invisible to the naked eye.
 
-\begin{figure}[h]
+\begin{figure}[ht]
  \centering
- \includegraphics[width=.75\textwidth]{figures/pdf/figure4.pdf}
+ \includegraphics[width=.85\textwidth]{figures/pdf/figure4.pdf}
  \caption{Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
-The blue pie charts convey a give minor allele frequency out of 100 percent, the green out 1 percent, and the red out of .01 percent.
-The data are from \citep{10002015global}.}
+The blue pie charts convey a given minor allele frequency out of 100 percent, the green out of 1 percent, and the red out of 0.01 percent.
+The data are from \citet{10002015global}.}
+\label{fig:freq}
 \end{figure}
 
-To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
-Specifically, we  use a “frequency scale” that is indicated in a legend below the map, and redundantly by using color, that indicates the scale for the frequencies displayed [Figure 4].
-Much like scientific notation, this allows a wide range of frequencies to be displayed (Table \ref{tab:rare}).
-
+  \begin{table}[h!]
     \begin{center}
-  \begin{table}[h]
     \begin{tabular}
@@
   \label{tab:rare}
+\end{center}
   \end{table}
 
-\end{center}
+To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
+Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:freq}].
+Much like scientific notation, this allows a wide range of frequencies to be displayed (Table \ref{tab:rare}).
 
@@
 
-To provide an interface to the population minor allele frequency data stored in the SQL database we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}.
+To provide an interface to the population minor allele frequency data, we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}.
 The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
-For the front-end, GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g.
-latitude, longitude, population label, sample size, and frequency scale.
+For the front-end, HTTP GET requests return json formatted allele frequency data and the meta-data associated with each population and genetic variant (e.g.~latitude, longitude, population label, sample size, and frequency scale).
 Genetic variants can be queried by chromosome position, rsid, or randomly.
-Example HTTP requests and json response can be seen in Examples 1,2,3 (below).
+Example HTTP requests and json response can be seen in the Appendix.
 
@@
 the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
-as determined in practice by the birthplaces of recent ancestors, such as grandparents).
+based on the birthplaces of recent ancestors, such as grandparents).
 We do not proscribe a general solution to this problem, and for the current defaults we use locations based on the approach taken in the source publications.
-A future feature will allow alternative location schema to be used for the populations in a dataset.
+A future feature will allow alternative location schemas to be used for the populations in a dataset.
 
-We also envision a variety of future extensions to the GGV that would allow for further dissection of geographical structure in large-scale population genomic datasets.
+We also envision a variety of future extensions to the GGV that would allow for further dissection of geographic structure in large-scale population genomic datasets.
 Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP.
-This feature would be relevant to both medical geneticists conducting genome-wide association studies with in interests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
-We imagine that incorporating a chromosomal browser such as jbrowse or a visualization inspired by the UCSC Genome Browser \citep{kent2002human} within the GGV would be greatly utilized by researchers and educators alike.
+This feature would be relevant to both medical geneticists conducting genome-wide association studies with interests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
+We imagine that incorporating a chromosomal browser such as jbrowse \citep{skinner2009jbrowse} within the GGV would be greatly utilized by researchers and educators alike.
 
@@
 
-Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank members of the Novembre Lab for supportive conversations.
+Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
+The content is soley the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.
+We acknowledge the Research Computer Center at the University of Chicago, especially H.~Birali Runesha, Jeff Tharsen, Richard Williams, and Alex Mueller, for on-going support and extensions of the GGV browser.
+We also thank John Zekos for web server administration and support.
+The authors would also like to thank members of the Novembre Lab for supportive conversations.
 
@@
 
-\subsection*{Example 1: Query by rsid}
+\bibliographystyle{genetics}
+\bibliography{paper}
+
+\section*{Appendix}
+\subsubsection*{Example 1: Query by rsid}
 
@@
     "rawfreq": 0.0752212389381
-  },
-  ...
+  }, ...
 ]
@@
 
-\subsection*{Example 2: Query by chromosome position}
+\subsubsection*{Example 2: Query by chromosome position}
 
@@
     "rawfreq": 0.0
-  }
-  ...
+  }, ...
 ]
@@
 
-\subsection*{Example 3: Random query}
+\subsubsection*{Example 3: Random query}
 
@@
     "rawfreq": 0.0
-  },
-  ...
+  }, ...
 ]
@@
 
-\clearpage
-\bibliographystyle{genetics}
-\bibliography{paper}
-
 \end{document}

### 2016-08-08 paper.tex
adjusting affiliation

@@
 \author[1]{Joseph H. Marcus$^{*}$}
-\author[1]{John Novembre\thanks{Address correspondence to JHM (\texttt{jhmarcus@uchicago.edu}) or JN (\texttt{jnovembre@uchicago.edu}).}}
+\author[1,2]{John Novembre\thanks{Address correspondence to JHM (\texttt{jhmarcus@uchicago.edu}) or JN (\texttt{jnovembre@uchicago.edu}).}}
 \affil[1]{\small{Department of Human Genetics, University of Chicago, Chicago, IL,USA}}
+\affil[2]{\small{Department of Ecology and Evolutionary Biology, University of Chicago, Chicago, IL,USA}}
 

### 2016-08-13 cabios-template/main.tex
started shortening and formartting for bioinformatics preprint

@@
 \title[short Title]{Visualizing the Geography of Genetic Variants}
-\author[Sample \textit{et~al}.]{Corresponding Author\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
-\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
-$^{\text{\sf 2}}$Department, Institution, City, Post Code,
-Country.}
+\author[Sample \textit{et~al}.]{Joseph H. Marcus\,$^{\text{\sfb 1}}$ and John Novembre\,$^{\text{\sfb 1, 2,}*}$}
+\address{$^{\text{\sf 1}}$Department of Human Genetics, University of Chicago, Chicago, 60637, US and \\
+$^{\text{\sf 2}}$Department of Ecology and Evolution, University of Chicago, Chicago, 60637, US.}
 
@@
 
-\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text
-Text Text Text Text Text.\\
-\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
-\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
-\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
+\abstract{One of the key characteristics of any genetic variant is its geographic distribution.
+The geographic distribution can shed light on where an allele first arose, what populations it has spread to, and in turn on how migration, genetic drift, and natural selection have acted.
+The distribution of a genetic variant can also be of great utility for medical/clinical geneticists.
+Collectively the geographic distribution of many genetic variants can reveal population structure.
+Here we develop an interactive visualization tool for rapidly displaying the geographic distribution of genetic variants.
+Through a REST API and dynamic front-end the \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}} browser provides maps of allele frequencies in populations distributed across the globe.\\
+\textbf{Contact:} \href{jnovembre@uchicago.edu}{jnovembre@uchicago.edu}\\
 \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
@@
 
-\section{Introduction}
+\section*{Introduction}
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text. Figure~\ref{fig:01}
-shows that the above method  Text Text Text Text  Text Text Text
-Text Text Text  Text Text.  \citep{Bag01} wants to know about
-{\ldots}{\ldots} text follows.
-\begin{equation}
-\sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
-\end{equation}
+Genetics researchers often face the problem that they have identified one or many genetic variants of interest using an approach such as a genome-wide association study and then would like to know the geographic distribution of the variant.
+For example, the researcher may hope to address: 1) implications for genomic medicine \citep[e.g.
+Is a risk allele geographically localized to a certain patient population?
+What population should be studied to observe variant carriers?][]{rosenberg2010genome}; or 2) the evolutionary history of the variant in question \citep[e.g.
+does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?][]{novembre2009spatial,coop2010using}.
+A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
 
-%\enlargethispage{12pt}
+Contemporary population genetics researchers are also faced with the challenge of large, high-dimensional datasets.
+For example, it is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
+One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \citep{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \citep{pritchard2000inference} or ADMIXTURE \citep{alexander2009fast}.
+While these approaches are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
+A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.
 
-\section{Approach}
+Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data-wrangling methods \citep{kandel2011research} and map-making techniques that are unfamiliar to most.
+Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.
 
-Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-.....
+Our work is inspired by past tools such as the ALFRED database \citep{rajeevan2011alfred} and the maps available on the HGDP Selection browser \citep{pickrell2009signals} whose allele frequency output and plots have been used in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been made available on the UCSC Genome Browser \citep[available under the HGDP Allele Freq track of the browser][]{kent2002human}.
 
-\begin{methods}
-\section{Methods}
+Taking advantage of recent advances in web-based visualization tools \citep{bostock2011d3}, we aim to address the significant visualization challenges that are inherit in the production of geographic allele frequency maps for large population genomic datasets \citep{10002015global}, \citep{lazaridis2014ancient}, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text
-text\vspace*{1pt}
+\section*{Approach}
 
-\begin{itemize}
-\item for bulleted list, use itemize
-\item for bulleted list, use itemize
-\item for bulleted list, use itemize\vspace*{1pt}
-\end{itemize}
+The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \citep{bostock2011d3} to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
+The front-end provides legends for the map and various configuration boxes to allow users to query different datasets or choose visualization options.
 
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-Text Text Text Text Text Text Text Text Text Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text\vadjust{\newpage}.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-Text Text Text Text Text Text Text Text Text Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\subsection{This is subheading}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text
-text. Text Text Text Text Text Text  Text Text Text Text Text Text
-Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\subsubsection{This is subsubheading}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text\break
-text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text
-textText Text Text Text Text Text  Text Text Text Text Text Text
-Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-\enlargethispage{6pt}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\begin{table}[!t]
-\processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}\toprule head1 &
-head2 & head3 & head4\\\midrule
-row1 & row1 & row1 & row1\\
-row2 & row2 & row2 & row2\\
-row3 & row3 & row3 & row3\\
-row4 & row4 & row4 & row4\\\botrule
-\end{tabular}}{This is a footnote}
-\end{table}
-
-\end{methods}
+In order to allow for easy access to commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global} or Human Genome Diversity project \citep{li2008worldwide}, we have developed a REST API \citep{grinberg2014flask} for accessing data.
+The API allows querying of allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs.
+While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
+We find this to be especially useful in teaching contexts, as it provides a highly visual way for learners to understand human genetic variation.
 
 \begin{figure}[!tpb]%figure1
-\fboxsep=0pt\colorbox{gray}{\begin{minipage}[t]{235pt} \vbox to 100pt{\vfill\hbox to
-235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}\vfill}
-\end{minipage}}
-%\centerline{\includegraphics{fig01.eps}}
-\caption{Caption, caption.}\label{fig:01}
+\centerline{\includegraphics[scale=.2]{figure1.png}}
+\caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
+Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:01}
+\label{fig:screenshot}
 \end{figure}
 
-%\begin{figure}[!tpb]%figure2
-%%\centerline{\includegraphics{fig02.eps}}
-%\caption{Caption, caption.}\label{fig:02}
-%\end{figure}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\subsection{Test1}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\section{Discussion}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text.
-
-Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text
-Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
-the above method Text Text. Text Text Text  Text Text Text Text
-Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
-method Text Text. Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-Text.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Conclusion}
-
-(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text
-Text Text Text Text Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
-\begin{enumerate}
-\item this is item, use enumerate
-\item this is item, use enumerate
-\item this is item, use enumerate
-\end{enumerate}
-
-Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
-above method  Text Text Text Text Text Text Text Text Text Text
-Text Text.  \citealp{Boffelli03} might want to know about text
-text text text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text\break Text.
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text
-Text Text  Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text\vspace*{-10pt}
+After a query, the GGV displays the allele frequencies for a set of populations as a collection of pie charts where each represents the minor and major allele frequency in a single population.
+Pie charts are displayed as points at a latitude and longitude assoicated with a population and the map boundaries are chosen based off of the geographic configuration of populations in a given dataset [Figure \ref{fig:screenshot}].
 
@@
 
-Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text
-text text text\vspace*{-12pt}
+We acknowledge the Research Computer Center at the University of Chicago, especially H.~Birali Runesha, Jeff Tharsen, Richard Williams, and Alex Mueller, for on-going support and extensions of the GGV browser.
+We also thank John Zekos for web server administration and support.
+The authors would also like to thank members of the Novembre Lab for supportive conversations.\\
+\textit{funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
+The content is soley the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.\\
 
-\section*{Funding}
-
-This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}
-
-%\bibliographystyle{natbib}
+\bibliographystyle{natbib}
 %\bibliographystyle{achemnat}
@@
 %\bibliographystyle{bioinformatics}
-%
 %\bibliographystyle{plain}
-%
-%\bibliography{Document}
 
-\begin{thebibliography}{}
+\bibliography{main}
 
-\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
-
-\bibitem[Bag {\it et~al}., 2001]{Bag01}
-Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
-
-\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-in nigral dopaminergic neurnol cell: correlation with the known
-pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-Res.}, \textbf{110}(Suppl. 1), 76--84.
-
-\bibitem[Lehmann, 1986]{Leh86}
-Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
-
-\bibitem[Crenshaw and Jones, 2003]{Cre03}
-Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-cancer management: one tumor, one chip. \textit{Bioinformatics},
-doi:10.1093/bioinformatics/btn000.
-
-\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-???--???.
-
-\bibitem[Bardet, 1920]{Bar20}
-Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-polydactylie et retinite pigmentaire (contribution a l'etude des
-formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-institution, Paris, France.
-
-\end{thebibliography}
 \end{document}

### 2016-08-14 cabios-template/main.tex
init bioinformatics draft done

@@
 
-\section*{Introduction}
+\section{Introduction}
 
@@
 One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \citep{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \citep{pritchard2000inference} or ADMIXTURE \citep{alexander2009fast}.
-While these approaches are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
+While these methods are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
 A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.
@@
 
-\section*{Approach}
+\section{Approach}
 
@@
 \begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[scale=.2]{figure1.png}}
+\centerline{\includegraphics[scale=.25]{figure1.png}}
 \caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
@@
 
+\section{Features}
+
+\textbf{Rare variants:} Representing the frequency of rare variants across populations can be difficult as allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}.
+To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
+Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:freq}].
+
+\textbf{Uncertainty in frequency data:}  One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
+To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:uncertainty}].
+
+\textbf{Overlapping populations:} In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
+To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:force}].
+Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
+
+\textbf{REST API:} To provide an interface to the population minor allele frequency data, we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}.
+The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
+Genetic variants can be queried by chromosome position, rsid, or randomly.
+
+\section{Conclusion}
+
+By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
+We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
+
 \section*{Acknowledgements}
@@
 The authors would also like to thank members of the Novembre Lab for supportive conversations.\\
-\textit{funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
+\textit{Funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
 The content is soley the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.\\

### 2016-08-16 cabios-template/main.tex
fixed typo
push before cleaning

@@
 The geographic distribution can shed light on where an allele first arose, what populations it has spread to, and in turn on how migration, genetic drift, and natural selection have acted.
-The distribution of a genetic variant can also be of great utility for medical/clinical geneticists.
-Collectively the geographic distribution of many genetic variants can reveal population structure.
+The geographic distribution of a genetic variant can also be of great utility for medical/clinical geneticists and collectively many genetic variants can reveal population structure.
 Here we develop an interactive visualization tool for rapidly displaying the geographic distribution of genetic variants.
@@
 
+\begin{figure*}[th!]%figure1
+\centerline{\includegraphics[width=0.7\textwidth]{figure1.jpg}}
+\caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
+Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:01}
+\label{fig:screenshot}
+\end{figure*}
+
 \section{Approach}
@@
 
-\begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[scale=.25]{figure1.png}}
-\caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
-Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:01}
-\label{fig:screenshot}
-\end{figure}
-
 After a query, the GGV displays the allele frequencies for a set of populations as a collection of pie charts where each represents the minor and major allele frequency in a single population.
@@
 To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
-Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:freq}].
+Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure 1,   Supplementary Table 1].
 
 \textbf{Uncertainty in frequency data:}  One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
-To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:uncertainty}].
+To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure 2].
 
 \textbf{Overlapping populations:} In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
-To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:force}].
+To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure 3].
 Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
@@
 The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
-Genetic variants can be queried by chromosome position, rsid, or randomly.
+Genetic variants can be queried by chromosome position, rsid, or randomly (see supplementary data for examples).
 
@@
 By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
-We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
+We emphasize that the overwhelming majority the human genome does not differ  across populations \citep[see a classic treatment in][]{lewontin1972apportionment}.
+We thus hope that the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in an approachable and intuitive form.
 
@@
 \textit{Funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
-The content is soley the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.\\
+The content is solely the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.\\
 

### 2016-08-16 supplement.tex
push before cleaning

--- 2016-08-08 paper.tex
@@
 
-\linenumbers
 \captionsetup{labelfont=bf}
 \frenchspacing
-\title{Visualizing the Geography of Genetic Variants}
+\title{\textbf{Supplementary Information}\\Visualizing the Geography of Genetic Variants}
 \author[1]{Joseph H. Marcus$^{*}$}
@@
 
-\begin{abstract}
-One of the key characteristics of any genetic variant is its geographic distribution.
-The geographic distribution can shed light on where an allele first arose, what populations it has spread to, and in turn on how migration, genetic drift, and natural selection have acted.
-The distribution of a genetic variant can also be of great utility for medical/clinical geneticists.
-Collectively the geographic distribution of many genetic variants can reveal population structure.
-As a result, visual inspection of geographic maps for genetic variants is common practice in genetic studies.
-Here we develop an interactive visualization tool for rapidly displaying the geographic distribution of genetic variants.
-Through a REST API and dynamic front-end the \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}} browser provides maps of allele frequencies in populations distributed across the globe.
-\end{abstract}
+\section*{Caveats and Future Directions}
 
-%%%%%%%%%%%%%%%%%%%%%%%
+A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.
+While doing so is generally immensely helpful, it has inherent complexity and limitations.
+For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g.
+the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
+based on the birthplaces of recent ancestors, such as grandparents).
+We do not proscribe a general solution to this problem, and for the current defaults we use locations based on the approach taken in the source publications.
+A future feature will allow alternative location schemas to be used for the populations in a dataset.
 
-\section*{Introduction}
+We also envision a variety of future extensions to the GGV that would allow for further dissection of geographic structure in large-scale population genomic datasets.
+Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP.
+This feature would be relevant to both medical geneticists conducting genome-wide association studies with interests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
+We imagine that incorporating a chromosomal browser such as jbrowse \citep{skinner2009jbrowse} within the GGV would be greatly utilized by researchers and educators alike.
 
-Genetics researchers often face the problem that they have identified one or many genetic variants of interest using an approach such as a genome-wide association study and then would like to know the geographic distribution of the variant.
-For example, the researcher may hope to address: 1) implications for genomic medicine \citep[e.g.
-Is a risk allele geographically localized to a certain patient population?
-What population should be studied to observe variant carriers?][]{rosenberg2010genome}; or 2) the evolutionary history of the variant in question \citep[e.g.
-does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?][]{novembre2009spatial,coop2010using}.
-A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
+\clearpage
 
-Contemporary population genetics researchers are also faced with the challenge of large, high-dimensional datasets.
-For example, it is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
-One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \citep{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \citep{pritchard2000inference} or ADMIXTURE \citep{alexander2009fast}.
-While these approaches are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
-A natural approach to gaining more insight to the overall structure of a population genetic dataset is to visually inspect what geographic patterns arise in allele frequency maps.
-
-Unfortunately, generating geographic allele frequency maps is time-consuming for the average researcher as it requires a combination of data-wrangling methods \citep{kandel2011research} and map-making techniques that are unfamiliar to most.
-Our aim here is to produce a tailored system for rapidly constructing informative geographic maps of allele frequency variation.
-
-Our work is inspired by past tools such as the ALFRED database \citep{rajeevan2011alfred} and the maps available on the HGDP Selection browser \citep{pickrell2009signals}.
-One of us (JN) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) \citep{wessel2013generic}, a powerful system of geographic plotting scripts for making static plots.
-The plots from the HGDP Selection Browser have proved useful, have appeared in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been made available on the UCSC Genome Browser \citep[available under the HGDP Allele Freq track of the browser][]{kent2002human}.
-
-Reference datasets for population genetic variation have greatly expanded since the release of the HGDP Illumina 650Y dataset \citep{li2008worldwide} that formed the basis of the HGDP Selection Browser maps.
-The most notable advance is the publication of the 1000 Genomes Phase 3 data \citep{10002015global} though additional datasets are continually coming online \citep[e.g.][]{lazaridis2014ancient}.
-In addition, novel approaches for data visualization have become more widely available.
-In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer useful methods for interactivity, the advantages of software development in modern web-browsers, a large open-source development community, and ease of sharing \citep{bostock2011d3}.
-
-%\begin{figure}[h]
-% \centering
-% \includegraphics[width=.4\textwidth]{figures/pdf/figure1.pdf}
-% \caption{Schematic diagram of the Geography of Genetic Variants (GGV) browser.
-We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genome Diversity Project (HGDP) and Population Reference Sample (POPRES).
-These allele frequencies are stored in a SQL database along side with geographic coordinates associated with each population.
-The front-end visualization was created using D3.js and accesses allele frequency and population meta-data from a REST API implemented using the python library flask.
-}
-%\end{figure}
-
-Taking advantage of these recent advances, we aim to address the significant visualization challenges that are inherit in the production of geographic allele frequency maps, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Fundamental Approach}
-
-The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \citep{bostock2011d3} to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
-The front-end provides legends for the map and various configuration boxes to allow users to query different datasets or choose visualization options.
-
-In order to allow for easy access to commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global} or Human Genome Diversity project \citep{li2008worldwide}, we have developed a REST API \citep{grinberg2014flask} for accessing data.
-The API allows querying of allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs.
-While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
-We find this to be especially useful in teaching contexts, as it provides a highly visual way for learners to understand human genetic variation.
-
-\begin{figure}[h]
- \centering
- \includegraphics[width=.95\textwidth]{figures/png/figure2.png}
- \caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
-Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}
-\label{fig:screenshot}
-\end{figure}
-
-After a query, the GGV displays the allele frequencies for a set of populations as a collection of pie charts where each represents the minor and major allele frequency in a single population.
-Pie charts are displayed as points at a latitude and longitude assoicated with a population and the map boundaries are chosen based off of the geographic configuration of populations in a given dataset [Figure \ref{fig:screenshot}].
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Representing uncertainty in frequency data}
-
-One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
-For some locations, sample sizes are small, and the reported allele frequency may be quite far from the true population frequency due to sampling error.
-To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:uncertainty}].
-
-\begin{figure}[h]
- \centering
- \includegraphics[width=.8\textwidth]{figures/png/figure3.png}
- \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
-The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.
-The frequency data and population identifiers are from \citet{novembre2008genes}.}
-\label{fig:uncertainty}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Representing rare variants in frequency data}
-
-An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset.
-This has not been a pervasive problem until recently, as most population genetic samples were genotyped on SNP arrays, which have been biased towards variants that are common in human populations (5-50 \% in minor allele frequency).
-With the combination of next generation sequencing technologies, new array designs focusing on rarer variants, and studies with thousands of individuals or more, it is now routine for the majority of variants to be rare \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}.
-In visualization schemes using proportional area to represent frequency (such as standard pie charts), rare variants would be represented as narrow slivers, nearly invisible to the naked eye.
+\section*{Figures and Tables}
 
@@
  \centering
- \includegraphics[width=.85\textwidth]{figures/pdf/figure4.pdf}
+ \includegraphics[width=.85\textwidth]{supp_figures/figure1.pdf}
  \caption{Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
@@
 
-To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
-Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:freq}].
-Much like scientific notation, this allows a wide range of frequencies to be displayed (Table \ref{tab:rare}).
+\begin{figure}[h]
+   \centering
+   \includegraphics[width=.8\textwidth]{supp_figures/figure2.pdf}
+   \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
+The transparency is scaled in proportion to the number of observed chromosomes in each population for a particular variant.
+The frequency data and population identifiers are from \citet{novembre2008genes}.}
+  \label{fig:uncertainty}
+\end{figure}
 
@@
  \centering
- \includegraphics[width=\textwidth]{figures/pdf/figure5.pdf}
+ \includegraphics[width=\textwidth]{supp_figures/figure3.pdf}
  \caption{Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional features of the interface}
-
-In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
-To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure 5].
-Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Access to the underlying frequency data}
-
-To provide an interface to the population minor allele frequency data, we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}.
-The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
-For the front-end, HTTP GET requests return json formatted allele frequency data and the meta-data associated with each population and genetic variant (e.g.~latitude, longitude, population label, sample size, and frequency scale).
-Genetic variants can be queried by chromosome position, rsid, or randomly.
-Example HTTP requests and json response can be seen in the Appendix.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Discussion}
-
-By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
-We also hope the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in a more approachable and intuitive form than alternative visualizations.
-
-A major challenge of using a geographic representation of genetic variation in humans is that the samples must be associated with a geographic location.
-While doing so is generally immensely helpful, it has inherent complexity and limitations.
-For example,  practitioners must make choices regarding representing where an individual was sampled for the study (e.g.
-the city of a major research center) or choosing a location that is more representative of an individual's ancestral origins (e.g.
-based on the birthplaces of recent ancestors, such as grandparents).
-We do not proscribe a general solution to this problem, and for the current defaults we use locations based on the approach taken in the source publications.
-A future feature will allow alternative location schemas to be used for the populations in a dataset.
-
-We also envision a variety of future extensions to the GGV that would allow for further dissection of geographic structure in large-scale population genomic datasets.
-Providing an interactive means of browsing neighboring variant sites near a SNP of interest would offer a unique view into patterns of linkage disequilibrium around that focal SNP.
-This feature would be relevant to both medical geneticists conducting genome-wide association studies with interests in fine mapping as well as population geneticists interested in scanning the genome to detect signatures of positive selection.
-We imagine that incorporating a chromosomal browser such as jbrowse \citep{skinner2009jbrowse} within the GGV would be greatly utilized by researchers and educators alike.
-
-%The constellation of common single-nucleotide polymorphism (SNP) alleles carried by an individual can provide some insight to the geographic origins of their recent ancestors.
-More recently, the increasing availability of data on rare variants is opening up new opportunities for ancestry inference.
-Rare variants are particularly informative for ancestry as they tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration events \citep{nelson2012abundance, mathieson2014demography}.
-By assessing the geographic distribution of rare variants carried by an individual one can gain information about their ancestry (e.g.
-\citep{BrownPasniuc2014PloSCompBio, KennewickManReport}).
-In these applications, geographic maps of the genetic variants an individual carries can be useful and informative.
-We imagine developing tools to upload personal genetic data from companies such as from 23andMe or AncestryDNA would allow users to visualize the set of rare variants they carry and could be of wide interest to the public.
-In order to achieve this goal we plan to extend the ability of the GGV API to query and return data from thousands of genetic variants at a time.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Acknowledgements}
-
-Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
-The content is soley the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.
-We acknowledge the Research Computer Center at the University of Chicago, especially H.~Birali Runesha, Jeff Tharsen, Richard Williams, and Alex Mueller, for on-going support and extensions of the GGV browser.
-We also thank John Zekos for web server administration and support.
-The authors would also like to thank members of the Novembre Lab for supportive conversations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{genetics}
-\bibliography{paper}
+\clearpage
 

### 2016-08-16 supplement/supplement.tex
added supplement and clean up main text

--- 2016-08-16 supplement.tex
@@
  \centering
- \includegraphics[width=.85\textwidth]{supp_figures/figure1.pdf}
+ \includegraphics[width=.85\textwidth]{figures/figure1.pdf}
  \caption{Example maps from the Geography of Genetic Variants browser displaying the use of frequency scales for more expressive representations of rare variation on geographic maps.
@@
         {lllll} \hline  \textbf{Sample Frequency}& \textbf{Display Frequency Scale} & \textbf{Displayed Image} \\
-        \hline 0.25 & 1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_blue.pdf}} \\
-        0.025 &  0.1 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_green.pdf}} \\
-        0.0025 &  0.01 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_red.pdf}} \\
-        0.00025 &  0.001 & \parbox[c]{1em}{\includegraphics[width=.3in]{./figures/pdf/pie_purple.pdf}}
+        \hline 0.25 & 1 & \parbox[c]{1em}{\includegraphics[width=.3in]{figures/pie_blue.pdf}} \\
+        0.025 &  0.1 & \parbox[c]{1em}{\includegraphics[width=.3in]{figures/pie_green.pdf}} \\
+        0.0025 &  0.01 & \parbox[c]{1em}{\includegraphics[width=.3in]{figures/pie_red.pdf}} \\
+        0.00025 &  0.001 & \parbox[c]{1em}{\includegraphics[width=.3in]{figures/pie_purple.pdf}}
     \end{tabular}
@@
    \centering
-   \includegraphics[width=.8\textwidth]{supp_figures/figure2.pdf}
+   \includegraphics[width=.8\textwidth]{figures/figure2.pdf}
    \caption{Example map from the Geography of Genetic Variants browser displaying the use of varying transparency of population pie charts to represent uncertainty in allele frequencies.
@@
  \centering
- \includegraphics[width=\textwidth]{supp_figures/figure3.pdf}
+ \includegraphics[width=\textwidth]{figures/figure3.pdf}
  \caption{Example maps from the Geography of Genetic Variants browser displaying the use of a force directed layout to limit visual clutter when many populations overlap in geographic position.
@@
 
+\bibliographystyle{natbib}
+\bibliography{supplement}
+
 \end{document}

### 2016-08-16 cabios-template/main.tex
added supplement and clean up main text
Edits to discussion; fixing supp fig/table number

@@
 To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
-Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure 1,   Supplementary Table 1].
+Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure S1,   Supplementary Table S1].
 
 \textbf{Uncertainty in frequency data:}  One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
-To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure 2].
+To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure S2].
 
 \textbf{Overlapping populations:} In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
-To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure 3].
+To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure S3].
 Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
@@
 The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
-Genetic variants can be queried by chromosome position, rsid, or randomly (see supplementary data for examples).
+Genetic variants can be queried by chromosome position, rsid, or randomly (see Supplementary appendix for examples).
 
@@
 By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists.
-We emphasize that the overwhelming majority the human genome does not differ  across populations \citep[see a classic treatment in][]{lewontin1972apportionment}.
-We thus hope that the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in an approachable and intuitive form.
+Also, for students of human diversity, it is often difficult to conceptualize classic statements regarding how most variation in humans is shared among populations \citep{lewontin1972apportionment} and how the fixation index $F_{ST}$ is relatively low globally \citep[10-15\%][]{10002015global}.
+We hope that the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in an approachable and intuitive form.
+One caveat of these maps is that one must interpret the geographic positioning of the plotted populations with care (see Supplementary Information for expanded discussion of this point).
 

### 2016-08-17 cabios-template/main.tex
final b4 submission
fixed funders

@@
 Through a REST API and dynamic front-end the \href{popgen.uchicago.edu/ggv}{\textit{Geography of Genetic Variants (GGV)}} browser provides maps of allele frequencies in populations distributed across the globe.\\
-\textbf{Contact:} \href{jnovembre@uchicago.edu}{jnovembre@uchicago.edu}\\
+\textbf{Contact:} \href{jhmarcus@uchicago.edu}{jhmarcus@uchicago.edu}, \href{jnovembre@uchicago.edu}{jnovembre@uchicago.edu}\\
 \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
@@
 The authors would also like to thank members of the Novembre Lab for supportive conversations.\\
-\textit{Funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute of General Medicine under training grant award number T32GM007197 for JHM.
+\textit{Funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute Of General Medical Sciences under training grant award number T32GM007197 for JHM.
 The content is solely the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.\\

### 2016-08-20 cabios-template/main.tex
trying to trim things up

@@
 
-\begin{figure*}[th!]%figure1
-\centerline{\includegraphics[width=0.7\textwidth]{figure1.jpg}}
+\section{Approach}
+
+\begin{figure}[!tpb]%figure1
+\centerline{\includegraphics[scale=0.2]{figure1.jpg}}
 \caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
-Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.}\label{fig:01}
-\label{fig:screenshot}
-\end{figure*}
-
-\section{Approach}
+Each pie chart represents a population with the blue slice of the pie displaying the frequency of the global minor allele.}\label{fig:01}\label{fig:screenshot}
+\end{figure}
 
@@
 The API allows querying of allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs.
-While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset.
-We find this to be especially useful in teaching contexts, as it provides a highly visual way for learners to understand human genetic variation.
-
 After a query, the GGV displays the allele frequencies for a set of populations as a collection of pie charts where each represents the minor and major allele frequency in a single population.
-Pie charts are displayed as points at a latitude and longitude assoicated with a population and the map boundaries are chosen based off of the geographic configuration of populations in a given dataset [Figure \ref{fig:screenshot}].
+Pie charts are displayed as points at a latitude and longitude associated with a population and the map boundaries are chosen based off of the geographic configuration of populations in a given dataset [Figure \ref{fig:screenshot}].
 
@@
 
-\textbf{Rare variants:} Representing the frequency of rare variants across populations can be difficult as allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}.
+\textbf{(1) Rare variants:} Representing the frequency of rare variants across populations can be difficult as allele frequencies between variants often differ greatly \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}.
 To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
 Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure S1,   Supplementary Table S1].
-
-\textbf{Uncertainty in frequency data:}  One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty.
-To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure S2].
-
-\textbf{Overlapping populations:} In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information.
-To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure S3].
+\textbf{(2) Uncertainty in frequency data:} We use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure S2].
+\textbf{(3) Overlapping populations:} We use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure S3].
 Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
-
-\textbf{REST API:} To provide an interface to the population minor allele frequency data, we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}.
-The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly.
-Genetic variants can be queried by chromosome position, rsid, or randomly (see Supplementary appendix for examples).
 
@@
 We hope that the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in an approachable and intuitive form.
-One caveat of these maps is that one must interpret the geographic positioning of the plotted populations with care (see Supplementary Information for expanded discussion of this point).
 
@@
 We also thank John Zekos for web server administration and support.
-The authors would also like to thank members of the Novembre Lab for supportive conversations.\\
-\textit{Funding}: Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute Of General Medical Sciences under training grant award number T32GM007197 for JHM.
-The content is solely the responsiblity of the authors and does not necessarily reflect the official view of the National Institutes of Health.\\
+The authors would also like to thank members of the Novembre Lab for supportive conversations.
+Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute Of General Medical Sciences under training grant award number T32GM007197 for JHM.
+The content is solely the responsibility of the authors and does not necessarily reflect the official view of the National Institutes of Health.
 

### 2016-08-24 cabios-template/main.tex
took out section header
remove conclusion header
revising with Joe

@@
 What population should be studied to observe variant carriers?][]{rosenberg2010genome}; or 2) the evolutionary history of the variant in question \citep[e.g.
-does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?][]{novembre2009spatial,coop2010using}.
+does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?][]{novembre2009spatial}.
 A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.
@@
 For example, it is not uncommon for a researcher in human genetics to have a dataset comprised of thousands of individuals measured at hundreds of thousands or even millions of single nucleotide variants (SNVs).
-One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \citep{price2006principal, patterson2006population}, or a model-based clustering method such as STRUCTURE \citep{pritchard2000inference} or ADMIXTURE \citep{alexander2009fast}.
+One common approach to visualizing such high-dimensional data is to compress the SNV dimensions down to a small number of latent factors, using a method such as principal components analysis \citep{patterson2006population}, or a model-based clustering method such as STRUCTURE \citep{pritchard2000inference}.
 While these methods are extremely valuable, researchers can use them too often without inspecting the underlying variant data in more detail.
@@
 
-Our work is inspired by past tools such as the ALFRED database \citep{rajeevan2011alfred} and the maps available on the HGDP Selection browser \citep{pickrell2009signals} whose allele frequency output and plots have been used in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been made available on the UCSC Genome Browser \citep[available under the HGDP Allele Freq track of the browser][]{kent2002human}.
+Our work is inspired by past tools such as the ALFRED database \citep{rajeevan2011alfred} and the maps available on the HGDP Selection browser \citep{pickrell2009signals} whose allele frequency output and plots have been used in research articles \citep[e.g.][]{pickrell2009signals,coop2009role}, books \citep[e.g.][]{dudley2013exploring}, and have been made available on the UCSC Genome Browser \citep[available under the HGDP Allele Freq track of the browser,][]{kent2002human}.
 
-Taking advantage of recent advances in web-based visualization tools \citep{bostock2011d3}, we aim to address the significant visualization challenges that are inherit in the production of geographic allele frequency maps for large population genomic datasets \citep{10002015global}, \citep{lazaridis2014ancient}, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
+Taking advantage of recent advances in web-based visualization tools \citep{bostock2011d3}, we aim to address the significant visualization challenges that are inherit in the production of geographic allele frequency maps for large population genomic datasets, including dynamic interaction, display of rare genetic variation, and representation of uncertainty in estimated allele frequencies due to variable sample sizes.
 
@@
 \begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[scale=0.2]{figure1.jpg}}
+\centerline{\includegraphics[scale=0.22]{figure1.jpg}}
 \caption{Example screenshot from the Geography of Genetic Variants browser using \citet{10002015global} data.
@@
 
-The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \citep{bostock2011d3} to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.
+The Geography of Genetic Variants browser (GGV) uses the scalable vector graphics and mapping utilities of D3.js \citep{bostock2011d3}.
 The front-end provides legends for the map and various configuration boxes to allow users to query different datasets or choose visualization options.
 
-In order to allow for easy access to commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global} or Human Genome Diversity project \citep{li2008worldwide}, we have developed a REST API \citep{grinberg2014flask} for accessing data.
-The API allows querying of allele frequencies by chromosome and position, by reference SNP identifier \citep{sherry2001dbsnp}, or randomly sampled SNPs.
-After a query, the GGV displays the allele frequencies for a set of populations as a collection of pie charts where each represents the minor and major allele frequency in a single population.
-Pie charts are displayed as points at a latitude and longitude associated with a population and the map boundaries are chosen based off of the geographic configuration of populations in a given dataset [Figure \ref{fig:screenshot}].
+In order to allow for easy access to commonly used public genomic datasets, such as the 1000 Genomes project \citep{10002015global} or Human Genome Diversity project \citep{li2008worldwide}, we have developed a REST API \citep{grinberg2014flask}.
+The API allows retreival of SNVs by position, rsid \citep{sherry2001dbsnp} or at random.
+After a query, the GGV displays the allele frequencies as a collection of pie charts where each represents the frequency of the globally minor allele in a single population [Figure \ref{fig:screenshot}].
 
-\section{Features}
-
-\textbf{(1) Rare variants:} Representing the frequency of rare variants across populations can be difficult as allele frequencies between variants often differ greatly \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}.
-To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible.
+We implement a variety of features: \textbf{(1) Rare variants.} Many alleles are rare \citep[e.g.][]{10002015global}, and displaying them can be a challenge with proportional scales that range from zero to one.
+To address this challenge we re-scale frequencies , so that small frequencies become visible.
 Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure S1,   Supplementary Table S1].
-\textbf{(2) Uncertainty in frequency data:} We use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure S2].
-\textbf{(3) Overlapping populations:} We use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure S3].
+\textbf{(2) Uncertainty in frequency data.} We use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure S2].
+\textbf{(3) Overlapping populations.} We use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure S3].
 Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.
-
-\section{Conclusion}
 
@@
 
-We acknowledge the Research Computer Center at the University of Chicago, especially H.~Birali Runesha, Jeff Tharsen, Richard Williams, and Alex Mueller, for on-going support and extensions of the GGV browser.
-We also thank John Zekos for web server administration and support.
-The authors would also like to thank members of the Novembre Lab for supportive conversations.
-Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0) to JN and the National Institute Of General Medical Sciences under training grant award number T32GM007197 for JHM.
-The content is solely the responsibility of the authors and does not necessarily reflect the official view of the National Institutes of Health.
+We acknowledge the Research Computer Center at the University of Chicago, especially Jeff Tharsen and Alex Mueller, for on-going support and development, as well as John Zekos for server administration support and members of the Novembre Lab.
+Support for this work was provided by the National Institutes of Health via the Big Data to Knowledge initiative (1U01 CA198933-0, JN) and the National Institute Of General Medical Sciences under training grant award number T32GM007197 (JHM).
 
